---
title: "Polinizadores de Orquídeas — Distribución Global"
author: "Natalia Palou y Caleb Pacheco"
date: "2025-11-18"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
runtime: shiny
---

<!--
  NOTE: this document uses `runtime: shiny` (see the YAML header above).
  Because of that it must be opened with **Run Document** in RStudio
  (the button replaces "Knit"), not with plain Knit. Plain Knit cannot
  start the Shiny app at the end.
-->

```{r setup, include=FALSE}
# include=FALSE -> this chunk runs but is not shown in the output.
# echo = TRUE      : show the code of every other chunk
# warning/message  : hide warnings and messages so the report stays clean
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## ¿Qué hicimos este semestre?

Este semestre estuvo dedicado a continuar buscando en los *papers* las
latitudes y longitudes en América para completar el Excel, y así seguir
avanzando en la recopilación y el análisis de la data.

El documento se organiza en: la preparación de los datos, un **mapa estático**
general, un conjunto de **mapas por subfamilia** y una **aplicación
interactiva (Shiny)** que permite filtrar el mapa por subfamilia. Todos los
mapas se construyen a partir de los mismos datos y de la misma función, de
modo que no hay código duplicado.

***

## Paquetes

```{r librerias}
library(readxl)   # read Excel files (.xlsx, .xls)
library(dplyr)    # data manipulation: filter, mutate, count, the %>% pipe
library(tidyr)    # tidyr::fill() -- carries a value down into blank rows
library(stringr)  # text/pattern helpers, mainly str_detect()
library(leaflet)  # interactive maps
library(shiny)    # the interactive app at the end
```

***

## Carga y preparación de datos

Leemos la hoja `species` del archivo Excel y limpiamos los nombres de las
columnas.

```{r carga-datos}
# Path to the source workbook. This is the ONLY place the file name lives:
# if the Excel is renamed, change just this line.
ruta_datos <- "Pollination_List_Thru_1 2024_merged26_08_13.xlsx"

# A workbook can hold several sheets; we explicitly read the "species" sheet.
data <- read_excel(ruta_datos, sheet = "species")

# Why clean the column names:
# the Excel headers have inconsistent capitalization ("Pollinator_type") and
# stray trailing spaces ("latitud "). We trim the spaces and lowercase
# everything once here, so the rest of the code can rely on tidy names.
colnames(data) <- tolower(trimws(colnames(data)))
```

Detectamos las columnas relevantes **por su nombre**, no por su posición. La
hoja crece y se reordena entre semestres; buscar por nombre evita que el
código se rompa si una columna cambia de lugar.

```{r detectar-columnas}
# Helper: given all column names and a search pattern, return the FIRST
# column name that matches, or NA if none matches.
#   - str_detect() tests each name against the pattern and returns TRUE/FALSE
#   - the pattern is a regular expression: "a|b" means "contains a OR b"
buscar_col <- function(nombres, patron) {
  encontradas <- nombres[str_detect(nombres, patron)]
  if (length(encontradas) == 0) NA_character_ else encontradas[1]
}

nombres  <- names(data)
lat_col  <- buscar_col(nombres, "lat")              # e.g. "latitud"
lon_col  <- buscar_col(nombres, "lon")              # e.g. "longitud"
poll_col <- buscar_col(nombres, "pollinator_type")  # preferred: the "type" column
if (is.na(poll_col)) poll_col <- buscar_col(nombres, "pollinator")  # fallback
loc_col  <- buscar_col(nombres, "locality|localidad|sitio|lugar")
spec_col <- buscar_col(nombres, "species|especie|taxon|orquidea|orquídea")
# "^genus$" matches the column named EXACTLY "genus". A loose "genus" pattern
# would also catch "genus count", which is a different column.
genus_col <- buscar_col(nombres, "^genus$")
subf_col  <- buscar_col(nombres, "^subfamily$")

# Coordinates are essential. If they are missing, stop now with a clear
# message instead of failing later with a confusing error.
if (is.na(lat_col) || is.na(lon_col)) {
  stop("No se encontraron columnas de latitud/longitud en la hoja 'species'.")
}

# Print which columns were picked, so it is easy to confirm the detection.
cat("Columna de latitud:    ", lat_col,   "\n")
cat("Columna de longitud:   ", lon_col,   "\n")
cat("Columna de polinizador:", poll_col,  "\n")
cat("Columna de género:     ", genus_col, "\n")
cat("Columna de subfamilia: ", subf_col,  "\n")
```

Construimos el conjunto de datos del mapa.

```{r preparar-datos}
# Coordinates arrive as TEXT, sometimes with stray spaces or a comma as the
# decimal separator ("18,234"). This helper turns that text into clean numbers.
a_numero <- function(x) {
  x <- gsub("\\s+", "", x)  # remove any spaces
  x <- gsub(",", ".", x)    # comma -> dot, because R needs a dot decimal
  suppressWarnings(as.numeric(x))  # text that is not a number becomes NA
}

# --- Carry "fill-down" columns into their blank rows ---
# In the Excel, `genus` and `subfamily` are written ONCE and left blank for
# the following rows that share the same value. (The `species` column, in
# turn, holds only the epithet, e.g. "nuda" -- so the full name is
# genus + epithet, "Apostasia nuda".)
# This helper turns the blank cells into NA and then carries the last value
# DOWN. It MUST run before filtering rows: a blank row borrows its value from
# the row above, so the original row order has to be intact.
rellenar_hacia_abajo <- function(df, col) {
  if (is.na(col)) return(df)                                # column not found
  v <- trimws(as.character(df[[col]]))
  df[[col]] <- ifelse(v == "" | is.na(v), NA_character_, v)  # blanks -> NA
  tidyr::fill(df, all_of(col), .direction = "down")          # carry down
}

data <- data %>%
  rellenar_hacia_abajo(genus_col) %>%
  rellenar_hacia_abajo(subf_col)

# About `.data[[lat_col]]` used below:
# lat_col is a *string* holding a column name (e.g. "latitud"). The `.data`
# pronoun lets us pick the column whose NAME is stored in that variable.
# Plain `data$lat_col` would NOT work -- it would look for a column literally
# called "lat_col".
data_mapa <- data %>%
  mutate(
    latitud  = a_numero(.data[[lat_col]]),
    longitud = a_numero(.data[[lon_col]]),

    # Standardize the pollinator label. The { } block is a small multi-line
    # expression: it runs top to bottom and the LAST line is the result
    # assigned to `polinizador`.
    polinizador = {
      v <- trimws(as.character(.data[[poll_col]]))
      v <- ifelse(is.na(v) | v == "", "No reportado", v)  # blanks -> label
      # Normalize spelling variants to a single form.
      # Add more recodes here as new inconsistencies appear in the Excel.
      v <- ifelse(tolower(v) == "wasp", "wasps", v)
      v
    },

    # Full scientific name = genus + epithet. The `genus` column was filled
    # down above; the `species` column holds only the epithet. paste() joins
    # them with a space; ifelse() avoids printing the text "NA" if either part
    # is missing.
    especie = {
      gen <- if (!is.na(genus_col)) as.character(.data[[genus_col]]) else NA_character_
      epi <- if (!is.na(spec_col))  as.character(.data[[spec_col]])  else NA_character_
      trimws(paste(ifelse(is.na(gen), "", gen), ifelse(is.na(epi), "", epi)))
    },
    localidad = if (!is.na(loc_col)) as.character(.data[[loc_col]]) else NA_character_,

    # Subfamily (already filled down above). Used further down to split the
    # data into one map per subfamily. Missing values get a clear label.
    subfamilia = {
      s <- if (!is.na(subf_col)) as.character(.data[[subf_col]]) else NA_character_
      ifelse(is.na(s) | s == "", "Sin subfamilia", s)
    },

    # Text shown in the popup when a point is clicked. <b>, <i> and <br> are
    # HTML (bold / italic / line break) -- leaflet renders the popup as a
    # small web page. Scientific names are shown in italics by convention.
    popup_info = paste0(
      "<b>Especie:</b> <i>", especie, "</i><br>",
      "<b>Localidad:</b> ",   localidad,   "<br>",
      "<b>Polinizador:</b> ", polinizador, "<br>",
      "<b>Lat:</b> ", round(latitud, 5),
      " | <b>Lon:</b> ", round(longitud, 5)
    )
  ) %>%
  # Keep only rows that ended up with usable coordinates. Most rows in the
  # Excel still have no coordinates -- that data entry is still in progress,
  # so a small number of mapped points here is expected, not a bug.
  filter(!is.na(latitud), !is.na(longitud))
```

### Resumen rápido

```{r resumen}
cat("Registros con coordenadas válidas:", nrow(data_mapa), "\n")
cat("Tipos de polinizador distintos:   ", dplyr::n_distinct(data_mapa$polinizador), "\n")
```

```{r tabla-polinizadores}
# How many mapped records each pollinator type has. knitr::kable() turns the
# small summary table into a nicely formatted table in the HTML output.
data_mapa %>%
  count(polinizador, name = "registros") %>%
  arrange(desc(registros)) %>%
  knitr::kable(caption = "Registros por tipo de polinizador")
```

***

## Mapa

Definimos una sola paleta de colores y una sola función, `construir_mapa()`,
que el mapa estático y la app de Shiny comparten. Así el mapa se describe en
un único lugar.

```{r mapa-funcion}
# All pollinator labels that exist, sorted -- this is the set of map layers.
dominio_colores <- sort(unique(data_mapa$polinizador))

colores_custom <- c(
  "#FF0000", "#FF8C00", "#FFD700", "#32CD32", "#00BFFF",
  "#0000FF", "#8A2BE2", "#FF1493", "#00FFFF", "#FF00FF"
)

# colorFactor() builds a function that maps each category (each pollinator
# type) to one color. `rep(..., length.out = n)` recycles the color list in
# case there are ever more pollinator types than colors defined above.
paleta_polinizador <- colorFactor(
  rep(colores_custom, length.out = length(dominio_colores)),
  domain = dominio_colores
)

# construir_mapa(): takes a prepared data frame and returns a Leaflet map.
# Defining it once means the static map and the Shiny app cannot drift apart.
construir_mapa <- function(df) {
  grupos <- sort(unique(df$polinizador))

  m <- leaflet(df) %>%
    addTiles() %>%                                  # standard street base map
    addProviderTiles(providers$Esri.WorldImagery)   # satellite imagery on top

  # We add ONE marker layer per pollinator type (instead of all points at
  # once). Each layer gets a name via `group = pol`; that name is what the
  # layer control below uses to switch a pollinator on/off.
  for (pol in grupos) {
    m <- m %>%
      addCircleMarkers(
        data = df %>% filter(polinizador == pol),
        # In leaflet, `~columna` means "use this column of the data".
        lng = ~longitud, lat = ~latitud,
        popup = ~popup_info,
        radius = 5, stroke = TRUE, weight = 1, fillOpacity = 0.7,
        color = ~paleta_polinizador(polinizador),  # color from the palette
        group = pol                                # layer name (see above)
      )
  }

  # The control box (top-right) that lets the user show/hide each layer.
  m %>%
    addLayersControl(
      overlayGroups = grupos,
      position = "topright",
      options = layersControlOptions(collapsed = FALSE)
    )
}
```

### Mapa estático

Cada punto representa una localidad con su tipo de polinizador reportado.

```{r mapa-estatico}
# Printing the leaflet object renders the map.
construir_mapa(data_mapa)
```

### Mapas por subfamilia

Un mapa estático por cada subfamilia que tiene registros georreferenciados.
Los colores siguen representando el tipo de polinizador, igual que en el mapa
general, de modo que los mapas son comparables entre sí.

```{r mapas-subfamilia}
# We build one static map per subfamily. A plain for-loop that prints each
# map does NOT render htmlwidgets correctly inside a single chunk, so instead
# we collect, for every subfamily, a heading + its map into an htmltools
# tagList and return that whole list -- knitr then renders every map in it.
# Subfamilies with no mapped points simply do not appear (sort(unique(...))
# only lists subfamilies present in data_mapa).
htmltools::tagList(
  lapply(sort(unique(data_mapa$subfamilia)), function(sf) {
    htmltools::tagList(
      htmltools::h4(sf),
      construir_mapa(filter(data_mapa, subfamilia == sf))
    )
  })
)
```

### Aplicación interactiva (Shiny)

La misma información, ahora dentro de una aplicación Shiny. Una app Shiny
tiene dos partes: la **UI** (lo que se ve) y el **server** (la lógica que
genera los resultados).

```{r app-shiny, echo=FALSE}
# echo = FALSE -> the app's code is not printed in the report, only the app.

# UI: the layout. It declares WHERE the controls and the map go, not how the
# map is built.
ui <- fluidPage(
  titlePanel("Mapa de polinizadores de orquídeas"),
  p("Cada punto representa una localidad con su tipo de polinizador reportado."),
  # Dropdown to focus the map on one subfamily, or "Todas" to show them all.
  selectInput(
    "subfamilia", "Subfamilia:",
    choices = c("Todas", sort(unique(data_mapa$subfamilia)))
  ),
  leafletOutput("mapa", height = "650px")  # placeholder, id = "mapa"
)

# server: the logic. renderLeaflet() builds the map and connects it to the
# placeholder whose id is "mapa". It re-runs whenever input$subfamilia changes.
server <- function(input, output, session) {
  output$mapa <- renderLeaflet({
    # Start from the full data; if a subfamily is chosen, keep only its rows.
    df <- data_mapa
    if (input$subfamilia != "Todas") {
      df <- filter(df, subfamilia == input$subfamilia)
    }
    construir_mapa(df)
  })
}

# shinyApp() ties the UI and the server together and launches the app.
shinyApp(ui = ui, server = server)
```

***

## ¿Qué nos falta?

Para el próximo semestre tenemos en agenda completar los datos de América y
comenzar con los datos de Asia, mientras resolvemos el problema del archivo
CSV actualizado para que sea el que se refleje correctamente en las gráficas.
Además, esperamos poder completar la integración utilizando el paquete Shiny
en Posit.
