---
title: "Orchid Pollinators — Global Distribution"
author: "Raymond L. Tremblay, Natalia Palou, Caleb Pacheco and Naan Lamoso"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-location: left
    theme: flatly
    df-print: paged
    code-fold: true          # code chunks start collapsed...
    code-summary: "Show code" # ...with this label to expand them
    code-tools: true         # global button to show/hide all code
server: shiny
lang: en
---

<!--
  This is the Quarto version of the original `mapa_final.rmd`. It uses Shiny
  (`server: shiny` in the header above), so run it with **Run Document** in
  RStudio, or with `quarto preview mapa_final.qmd` in the terminal — not a plain
  static render, which cannot start the Shiny app at the end.

  Data source: the ENRICHED workbook produced by `pollinators_wrangling.qmd`
  (sheet `species` = the records + parsed poll_* columns, and `pollinators_long`
  = one pollinator taxon per row). Run that document first. The map uses the
  `species` sheet; the pollinator drill-down (group > family > genus) uses
  `pollinators_long`.

  The objects the app needs live in `#| context: setup` cells: they run once at
  startup and are available to both the static maps and the Shiny server.
-->

```{r}
#| label: setup
#| context: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

Orchidaceae is among the largest families of flowering plants, and its
pollination systems are correspondingly diverse. Records of who pollinates what,
however, are scattered across a century of primary literature and are rarely
tied to explicit coordinates, which makes the geography of orchid pollination
difficult to examine directly.

This document maps the georeferenced portion of a global compilation of orchid
pollination records. The underlying database was assembled by Ackerman *et al.*
(2023) and covers roughly 3,100 species; the work presented here adds
coordinates to those records, species by species, from the localities reported
in the source publications.

Georeferencing is in progress and proceeds by region, the Americas first and
Asia to follow. The maps below therefore show the records resolved to date
rather than the full database, and coverage is uneven between subfamilies.

The document presents a map of all georeferenced records, a set of maps by
subfamily, and an interactive application that filters by subfamily and resolves
the pollinators from group to family to genus. All three are built from one
dataset and one mapping function.

### Data source

Pollination records are drawn from the global orchid pollination database of
Ackerman JD, Phillips RD, Tremblay RL, Karremans A, Reiter N, Peter CI, Bogarín
D, Pérez-Escobar OA & Liu H (2023), *Beyond the various contrivances by which
orchids are pollinated: global patterns in orchid pollination biology*,
**Botanical Journal of the Linnean Society** 202: 295–324, archived at
<https://doi.org/10.5281/zenodo.7263689> under CC BY 4.0.

***

## Packages

```{r}
#| label: packages
#| context: setup
library(readxl)   # read Excel files (.xlsx, .xls)
library(openxlsx) # read the enriched workbook's sheets
library(dplyr)    # data manipulation: filter, mutate, count, the %>% pipe
library(tidyr)    # tidyr::fill() -- carries a value down into blank rows
library(stringr)  # text/pattern helpers, mainly str_detect()
library(ggplot2)  # the coverage / composition figure
library(leaflet)  # interactive maps
library(shiny)    # the interactive app at the end
```

***

## Loading and preparing the data

The `species` sheet of the enriched workbook is read and its column names are
normalised.

```{r}
#| label: load-data
#| context: setup
# Path to the enriched workbook. This is the ONLY place the file name lives:
# if it is renamed, change just this line.
data_path <- "Pollination_List_pollinators_enriched.xlsx"

# The `species` sheet drives the map; `pollinators_long` drives the drill-down.
data      <- read.xlsx(data_path, sheet = "species")
poll_long <- read.xlsx(data_path, sheet = "pollinators_long")

# Why clean the column names:
# the headers can have inconsistent capitalization and stray trailing spaces.
# We trim and lowercase once here so the rest of the code relies on tidy names.
colnames(data) <- tolower(trimws(colnames(data)))

# A stable row id per record, matching pollinators_long$row (same sheet order).
data$row <- seq_len(nrow(data))
```

Columns are identified **by name rather than by position**. The workbook grows
and is periodically reordered, so matching on name keeps the code stable when a
column moves.

```{r}
#| label: detect-columns
#| context: setup
# Helper: given all column names and a search pattern, return the FIRST
# column name that matches, or NA if none matches.
find_col <- function(col_names, pattern) {
  found <- col_names[str_detect(col_names, pattern)]
  if (length(found) == 0) NA_character_ else found[1]
}

col_names <- names(data)
lat_col   <- find_col(col_names, "lat")              # e.g. "latitud"
lon_col   <- find_col(col_names, "lon")              # e.g. "longitud"
# Two pollinator columns, and the order matters -- see the note in the next
# chunk. `poll_groups` is the one derived by the wrangling doc from the
# free-text `pollinators` cell; `pollinator_type` is the older hand-typed column.
grp_col   <- find_col(col_names, "^poll_groups$")    # preferred: the parsed groups
poll_col  <- find_col(col_names, "^pollinator_type$")# fallback: the legacy column
if (is.na(poll_col)) poll_col <- find_col(col_names, "pollinator_type")
loc_col   <- find_col(col_names, "locality|localidad|sitio|lugar")
spec_col  <- find_col(col_names, "^species$|especie|taxon")
# "^genus$" matches the column named EXACTLY "genus". A loose "genus" pattern
# would also catch "genus count", which is a different column.
genus_col <- find_col(col_names, "^genus$")
subf_col  <- find_col(col_names, "^subfamily$")

# Coordinates are essential. If they are missing, stop now with a clear message.
if (is.na(lat_col) || is.na(lon_col)) {
  stop("No latitude/longitude columns were found in the 'species' sheet.")
}

cat("Latitude column:  ", lat_col,   "\n")
cat("Longitude column: ", lon_col,   "\n")
cat("Pollinator group column (preferred):", grp_col,  "\n")
cat("Pollinator type column (fallback):  ", poll_col, "\n")
cat("Genus column:     ", genus_col, "\n")
cat("Subfamily column: ", subf_col,  "\n")
```

```{r}
#| label: prepare-helpers
#| context: setup
# Coordinates arrive as TEXT, sometimes with stray spaces or a comma as the
# decimal separator ("18,234"). This helper turns that text into clean numbers.
to_number <- function(x) {
  x <- gsub("\\s+", "", x)  # remove any spaces
  x <- gsub(",", ".", x)    # comma -> dot, because R needs a dot decimal
  suppressWarnings(as.numeric(x))
}

# Carry `genus`/`subfamily` down into their blank rows (idempotent if already
# filled). MUST run before filtering, so the original row order stays intact.
fill_down <- function(df, col) {
  if (is.na(col)) return(df)
  v <- trimws(as.character(df[[col]]))
  df[[col]] <- ifelse(v == "" | is.na(v), NA_character_, v)
  tidyr::fill(df, all_of(col), .direction = "down")
}

data <- data %>% fill_down(genus_col) %>% fill_down(subf_col)
```

The map dataset is then assembled: numeric coordinates, the pollinator group
that determines each point's colour, and the labels used in the popups.

Two columns record the pollinator, and the choice between them matters.
`pollinator_type` is a legacy field populated for only 99 of the 3,161 records
(64 of the georeferenced ones). Relying on it alone leaves 107 of the mapped
points unattributed, although 65 of those records name their pollinators to
species in the free-text `pollinators` field. `poll_groups`, derived from that
free text during data preparation, resolves 126 of the georeferenced records and
is therefore read first, with `pollinator_type` retained as a fallback.

```{r}
#| label: prepare-map-data
#| context: setup
map_data <- data %>%
  mutate(
    lat = to_number(.data[[lat_col]]),
    lon = to_number(.data[[lon_col]]),

    # Every group recorded for this record, e.g. "Bees; Diptera". See above for
    # why this reads `poll_groups` first and `pollinator_type` only as a fallback.
    pollinator_all = {
      g <- if (!is.na(grp_col)) trimws(as.character(.data[[grp_col]])) else NA_character_
      ifelse(is.na(g) | g %in% c("", "NA"), NA_character_, g)
    },

    # The single group used for the point colour and its map layer. A record
    # with several groups is plotted under the FIRST one, so every point belongs
    # to exactly one layer; the popup still lists all of them.
    pollinator = {
      primary <- sub(";.*$", "", pollinator_all)

      # Fallback: the legacy column, standardized to full taxonomic names.
      v <- tolower(trimws(as.character(.data[[poll_col]])))
      recode_poll <- c(bees = "Bees", diptera = "Diptera", wasp = "Wasps",
                       wasps = "Wasps", butterfly = "Lepidoptera",
                       lepid = "Lepidoptera", coleo = "Coleoptera")
      legacy <- unname(recode_poll[v])                                             # known labels
      legacy <- ifelse(is.na(legacy) & !is.na(v) & nzchar(v), str_to_title(v), legacy)

      out <- ifelse(!is.na(primary), primary, legacy)
      # "Not reported" now means what it says: no pollinator recorded anywhere.
      ifelse(is.na(out), "Not reported", out)
    },

    # Full scientific name = genus + epithet (genus was filled down above;
    # the `species` column holds only the epithet).
    sci_name = {
      gen <- if (!is.na(genus_col)) as.character(.data[[genus_col]]) else NA_character_
      epi <- if (!is.na(spec_col))  as.character(.data[[spec_col]])  else NA_character_
      trimws(paste(ifelse(is.na(gen), "", gen), ifelse(is.na(epi), "", epi)))
    },
    locality = if (!is.na(loc_col)) as.character(.data[[loc_col]]) else NA_character_,

    subfamily_label = {
      s <- if (!is.na(subf_col)) as.character(.data[[subf_col]]) else NA_character_
      ifelse(is.na(s) | s == "", "No subfamily", s)
    }
  ) %>%
  filter(!is.na(lat), !is.na(lon))

```

The named pollinator taxa for each record are collapsed into a single block per
record, grouped by family.

```{r}
#| label: prepare-pollinator-names
#| context: setup
# Keep only the pollinator rows that belong to a georeferenced (mapped) record,
# so the drill-down menus only offer pollinators that actually appear on the map.
poll_long <- poll_long %>% filter(row %in% map_data$row)

# For each mapped record, an HTML block listing its pollinator SPECIES, grouped
# by family, so the popup shows the named pollinators when available.
poll_popup <- poll_long %>%
  mutate(name = case_when(
    rank == "taxon" ~ paste0("<i>", poll_taxon, "</i>"),
    rank == "genus" ~ paste0("<i>", poll_genus, "</i> sp."),
    TRUE            ~ verbatim)) %>%
  group_by(row, poll_family) %>%
  summarise(line = paste0(ifelse(is.na(poll_family[1]), "", paste0(poll_family[1], ": ")),
                          paste(unique(name), collapse = ", ")), .groups = "drop_last") %>%
  summarise(poll_species = paste(line, collapse = "<br>"), .groups = "drop")

map_data <- map_data %>% left_join(poll_popup, by = "row")

```

Finally, the popup content itself.

```{r}
#| label: prepare-popup
#| context: setup
# Build the popup. A trait row is shown only when that trait is recorded.
trait_row <- function(label, value) ifelse(is.na(value), "",
  paste0("<b>", label, ":</b> ", value, "<br>"))

map_data <- map_data %>%
  mutate(popup_info = paste0(
    "<div style='font-size:13px;max-width:330px'>",
    "<div style='font-size:15px;margin-bottom:2px'><b><i>", sci_name, "</i></b></div>",
    "<b>Subfamily:</b> ", subfamily_label, "<br>",
    "<b>Locality:</b> ", ifelse(is.na(locality), "—", locality), "<br>",
    "<b>Coordinates:</b> ", round(lat, 4), " , ", round(lon, 4), "<br>",
    trait_row("Growth form",     growth_form),
    trait_row("Climate",         climate),
    trait_row("Compatibility",   compatibility),
    trait_row("Breeding system", breeding_system),
    trait_row("Reward",          reward),
    trait_row("Deception",       deception),
    ifelse(is.na(mycoheterotrophic), "", paste0("<b>", mycoheterotrophic, "</b><br>")),
    # Show the full group list only when there is more than one, since a single
    # group is already obvious from the point's colour.
    ifelse(is.na(pollinator_all) | !grepl(";", pollinator_all), "",
           paste0("<b>Pollinator groups:</b> ", gsub("; *", ", ", pollinator_all), "<br>")),
    "<b>Pollinators:</b> ",
    ifelse(is.na(poll_species) | poll_species == "",
           paste0(pollinator, " <span style='color:#888'>(no species recorded)</span>"),
           paste0("<br>", poll_species)),
    "</div>"
  ))
```

***

## Mapping

We define a single colour palette and a single function, `build_map()`, shared
by the static map and the Shiny app, so the map is described in just one place.

```{r}
#| label: map-function
#| context: setup
# Fixed colour per pollinator group (colourblind-friendly Okabe-Ito), the SAME
# palette as pollinator_map.qmd so both maps look alike. Each group's colour is
# used for its map points AND for the coloured dot on its layer-control
# checkbox, so the checkboxes double as the legend (no separate legend box).
group_colors <- c(Bees = "#E69F00", Diptera = "#56B4E9", Wasps = "#D55E00",
                  Coleoptera = "#009E73", Lepidoptera = "#CC79A7", Aves = "#0072B2",
                  Hemiptera = "#F0E442", "Hymenoptera (unassigned)" = "#8C6D31",
                  "Not reported" = "#BBBBBB")
color_of <- function(g) unname(ifelse(g %in% names(group_colors), group_colors[g], "#999999"))

```

`build_map()` converts a prepared data frame into a Leaflet map.

```{r}
#| label: build-map-function
#| context: setup
# build_map(): takes a prepared data frame and returns a Leaflet map. Defining
# it once means the static map and the Shiny app cannot drift apart.
build_map <- function(df) {
  groups <- sort(unique(df$pollinator))

  # Each layer's label is a coloured dot (same colour as its points) followed
  # by the group name; used as BOTH the marker group and the checkbox label.
  swatch <- function(g) sprintf(
    "<span style='display:inline-block;width:11px;height:11px;border-radius:50%%;background:%s;border:1px solid #444;margin-right:5px;vertical-align:middle'></span>%s",
    color_of(g), g)
  labels <- setNames(vapply(groups, swatch, character(1)), groups)

  m <- leaflet(df) %>%
    addTiles() %>%                                  # standard street base map
    addProviderTiles(providers$Esri.WorldImagery)   # satellite imagery on top

  # One marker layer per pollinator group, each in its own colour. The layer
  # name (`group=`) is the coloured HTML label, which the layer control renders.
  for (pol in groups) {
    m <- m %>%
      addCircleMarkers(
        data = df %>% filter(pollinator == pol),
        lng = ~lon, lat = ~lat,
        popup = ~popup_info,
        radius = 5, stroke = TRUE, weight = 1, fillOpacity = 0.7,
        color = color_of(pol),
        group = labels[[pol]]
      )
  }

  # Frame the map explicitly. Leaflet's automatic fit zooms to the maximum level
  # when a layer holds a single point, dropping the reader into a street-level
  # view with no geographic context. We therefore widen the bounding box to a
  # minimum visible extent before fitting.
  min_km  <- 10                                   # smallest extent to display
  deg_lat <- min_km / 111                         # 1 degree of latitude ~ 111 km
  lat_rng <- range(df$lat, na.rm = TRUE)
  lon_rng <- range(df$lon, na.rm = TRUE)

  # A degree of longitude shrinks with latitude, so scale it by cos(latitude)
  # to keep the window roughly square on the ground. The floor of 0.2 keeps the
  # correction finite near the poles.
  deg_lon <- min_km / (111 * max(cos(mean(lat_rng) * pi / 180), 0.2))

  widen <- function(r, min_span) {
    if (is.finite(diff(r)) && diff(r) >= min_span) return(r)
    mid <- mean(r)
    c(mid - min_span / 2, mid + min_span / 2)
  }
  lat_rng <- widen(lat_rng, deg_lat)
  lon_rng <- widen(lon_rng, deg_lon)

  m <- m %>% fitBounds(lon_rng[1], lat_rng[1], lon_rng[2], lat_rng[2])

  # The control box (top-right): a coloured, toggleable checkbox per group.
  m %>%
    addLayersControl(
      overlayGroups = unname(labels[groups]),
      position = "topright",
      options = layersControlOptions(collapsed = FALSE)
    )
}
```

## Distribution of all georeferenced records

Each point marks a locality at which pollination was observed, coloured by the
pollinator group recorded there.

```{r}
#| label: static-map
build_map(map_data)
```

## Maps by subfamily

One static map per subfamily that has georeferenced records. The colours still
represent the pollinator type, just like the general map, so the maps are
comparable with one another.

```{r}
#| label: subfamily-maps
htmltools::tagList(
  lapply(sort(unique(map_data$subfamily_label)), function(sf) {
    htmltools::tagList(
      htmltools::h4(sf),
      build_map(filter(map_data, subfamily_label == sf))
    )
  })
)
```

## Coverage and pollinator composition

Georeferencing is uneven across the family. The figure below shows, for each
subfamily, the proportion of its records that now carry coordinates, with the
counts given alongside.

```{r}
#| label: coverage-figure
#| fig-width: 7.2
#| fig-height: 2.5
#| fig-alt: "Horizontal bar chart of georeferencing coverage by orchid subfamily. Each bar shows the percentage of that subfamily's records that carry coordinates, labelled with the count of georeferenced records over the total held."
valid_sf <- c("Apostasioideae", "Vanilloideae", "Cypripedioideae",
              "Orchidoideae", "Epidendroideae")

coverage <- data %>%
  mutate(sf = trimws(as.character(.data[[subf_col]]))) %>%
  filter(sf %in% valid_sf) %>%
  group_by(subfamily = sf) %>%
  summarise(in_database  = dplyr::n(),
            georeferenced = sum(row %in% map_data$row),
            .groups = "drop") %>%
  mutate(pct = 100 * georeferenced / in_database) %>%
  arrange(georeferenced) %>%
  mutate(subfamily = factor(subfamily, levels = subfamily),
         label = paste0(format(georeferenced, big.mark = ",", trim = TRUE), " / ",
                        format(in_database,   big.mark = ",", trim = TRUE)))

# Thin bars with rounded ends, drawn as round-capped segments: a recessive grey
# track for the whole, a solid bar for the part achieved. These three constants
# are reused by the pollinator figure below, so the two read as one system.
bar    <- 3.4          # bar thickness, points
track  <- "#E2E2DD"    # recessive track
accent <- "#0072B2"    # single hue: this figure encodes magnitude, not identity

ggplot(coverage, aes(y = subfamily)) +
  geom_segment(aes(x = 0, xend = 100, yend = subfamily),
               colour = track, linewidth = bar, lineend = "round") +
  # Subfamilies with no coordinates yet show an empty track rather than a stub.
  geom_segment(data = ~ subset(.x, georeferenced > 0),
               aes(x = 0, xend = pmax(pct, 0.8), yend = subfamily),
               colour = accent, linewidth = bar, lineend = "round") +
  geom_text(aes(x = 104, label = label), hjust = 0, size = 3.2, colour = "grey30") +
  scale_x_continuous(limits = c(0, 132), expand = c(0, 0)) +
  labs(x = NULL, y = NULL,
       caption = "Bar length is the percentage of that subfamily's records carrying coordinates; the figures give georeferenced records over records held.") +
  theme_minimal(base_size = 12) +
  theme(panel.grid    = element_blank(),
        axis.text.x   = element_blank(),
        axis.ticks    = element_blank(),
        plot.caption  = element_text(hjust = 0, colour = "grey45", size = 8.5,
                                     margin = margin(t = 10)),
        plot.margin   = margin(4, 4, 4, 4))
```

Composition of the mapped records by pollinator group follows, in the same
colours as the map points. Records with more than one recorded group are counted
under the first; the map popups list every group for a record. *Not reported*
means no pollinator is recorded anywhere in that row, and *Hymenoptera
(unassigned)* marks records whose free-text entry names an order but neither a
family nor a genus the parser could resolve.

```{r}
#| label: pollinator-composition
#| fig-width: 7.2
#| fig-height: 3.4
#| fig-alt: "Horizontal bar chart of georeferenced orchid pollination records by pollinator group, ordered from most to fewest records. Each bar is labelled with its count and coloured with the same scheme used for the map points."
comp <- map_data %>%
  count(pollinator, name = "records") %>%
  arrange(records) %>%
  mutate(pollinator = factor(pollinator, levels = pollinator))

ggplot(comp, aes(y = pollinator)) +
  geom_segment(aes(x = 0, xend = max(records), yend = pollinator),
               colour = track, linewidth = bar, lineend = "round") +
  geom_segment(aes(x = 0, xend = records, yend = pollinator, colour = pollinator),
               linewidth = bar, lineend = "round") +
  # The count sits at the end of every bar, so the figure never relies on
  # colour alone to be read.
  geom_text(aes(x = max(records) * 1.05, label = records),
            hjust = 0, size = 3.2, colour = "grey30") +
  scale_colour_manual(
    values = setNames(color_of(levels(comp$pollinator)), levels(comp$pollinator)),
    guide  = "none") +
  scale_x_continuous(limits = c(0, max(comp$records) * 1.16), expand = c(0, 0)) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid  = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks  = element_blank(),
        plot.margin = margin(4, 4, 4, 4))
```

***

## Interactive exploration

The same records, filterable. Selecting a **subfamily** restricts the map;
selecting a pollinator **group** narrows the **family** menu, which in turn
narrows the **genus** menu. The controls combine, and *All* clears a level.

```{r}
#| label: app-ui
fluidRow(
  column(3, selectInput("subfamily", "Subfamily:",
                        choices = c("All", sort(unique(map_data$subfamily_label))))),
  column(3, selectInput("pgroup", "Pollinator group:",
                        choices = c("All", sort(unique(na.omit(poll_long$poll_group)))))),
  # Family and genus start already populated with every present value, and are
  # narrowed by the selection(s) above them (see the server cascade).
  column(3, selectInput("pfamily", "Pollinator family:",
                        choices = c("All", sort(unique(na.omit(poll_long$poll_family)))))),
  column(3, selectInput("pgenus",  "Pollinator genus:",
                        choices = c("All", sort(unique(na.omit(poll_long$poll_genus))))))
)
leafletOutput("map", height = "650px")
```

```{r}
#| label: app-server
#| context: server

# --- Cascade: subfamily -> group -> family -> genus ------------------------
# Each pollinator menu offers ONLY values that actually occur among the mapped
# records of the current subfamily and the selection(s) above it. So, for
# example, Cypripedioideae never lists Aves, and the genus menu drops any genus
# with no record for the chosen subfamily + group + family.

# Mapped pollinator rows restricted to the chosen subfamily.
poll_base <- reactive({
  pl <- poll_long
  if (input$subfamily != "All")
    pl <- pl %>% filter(row %in% map_data$row[map_data$subfamily_label == input$subfamily])
  pl
})

```

```{r}
#| label: app-server-cascade
#| context: server
# Subfamily changes -> refill group, family and genus for that subfamily.
observeEvent(input$subfamily, {
  pl <- poll_base()
  updateSelectInput(session, "pgroup",
    choices = c("All", sort(unique(na.omit(pl$poll_group)))),  selected = "All")
  updateSelectInput(session, "pfamily",
    choices = c("All", sort(unique(na.omit(pl$poll_family)))), selected = "All")
  updateSelectInput(session, "pgenus",
    choices = c("All", sort(unique(na.omit(pl$poll_genus)))),  selected = "All")
}, ignoreInit = TRUE)

# Group changes -> refill family + genus (within subfamily + group).
observeEvent(input$pgroup, {
  pl <- poll_base()
  if (input$pgroup != "All") pl <- pl %>% filter(poll_group == input$pgroup)
  updateSelectInput(session, "pfamily",
    choices = c("All", sort(unique(na.omit(pl$poll_family)))), selected = "All")
  updateSelectInput(session, "pgenus",
    choices = c("All", sort(unique(na.omit(pl$poll_genus)))),  selected = "All")
}, ignoreInit = TRUE)

# Family changes -> refill genus (within subfamily + group + family).
observeEvent(input$pfamily, {
  pl <- poll_base()
  if (input$pgroup  != "All") pl <- pl %>% filter(poll_group  == input$pgroup)
  if (input$pfamily != "All") pl <- pl %>% filter(poll_family == input$pfamily)
  updateSelectInput(session, "pgenus",
    choices = c("All", sort(unique(na.omit(pl$poll_genus)))), selected = "All")
}, ignoreInit = TRUE)

```

```{r}
#| label: app-server-map
#| context: server
# --- Which mapped records match the pollinator drill-down -------------------
sel_rows <- reactive({
  pl <- poll_long
  if (input$pgroup  != "All") pl <- pl %>% filter(poll_group  == input$pgroup)
  if (input$pfamily != "All") pl <- pl %>% filter(poll_family == input$pfamily)
  if (input$pgenus  != "All") pl <- pl %>% filter(poll_genus  == input$pgenus)
  unique(pl$row)
})

output$map <- renderLeaflet({
  df <- map_data
  if (input$subfamily != "All") df <- filter(df, subfamily_label == input$subfamily)
  # Apply the pollinator filter only when at least one level is set, so that
  # with no drill-down the "Not reported" points (absent from poll_long) still show.
  if (input$pgroup != "All" || input$pfamily != "All" || input$pgenus != "All") {
    df <- filter(df, row %in% sel_rows())
  }
  build_map(df)
})
```

***

