---
title: "Orchid Pollinators — Global Distribution"
author: "Natalia Palou and Caleb Pacheco"
date: "2025-11-18"
format:
  html:
    toc: true
    toc-location: left
    theme: flatly
    df-print: paged
    code-fold: true          # code chunks start collapsed...
    code-summary: "Show code" # ...with this label to expand them
    code-tools: true         # global button to show/hide all code
    embed-resources: true    # inline every JS/CSS asset -> ONE portable .html
lang: en
---

<!--
  STATIC, SELF-CONTAINED version of `mapa_final.qmd`, for sharing by email.

  Differences from `mapa_final.qmd`:
    * no `server: shiny` -> renders with a plain `quarto render`, no live R
      session needed to view it;
    * `embed-resources: true` -> every JS/CSS asset is inlined, so the output
      is ONE .html file with no `_files/` folder beside it;
    * the Shiny drill-down app at the end is removed (it cannot exist in a
      static file). The static map and the per-subfamily maps are unchanged and
      fully interactive: pan, zoom, click a point for its popup, and toggle
      pollinator groups with the checkboxes.

  Render with:  quarto render mapa_final_static.qmd
  Then email `mapa_final_static.html` on its own. Do NOT send the workbook --
  the data is baked into the file.

  NOTE: the base map tiles (OpenStreetMap + Esri satellite) stream from the web,
  so the recipient needs an internet connection when they open it. Everything
  else -- points, popups, data -- is inside the file.

  Data source: the ENRICHED workbook produced by `pollinators_wrangling.qmd`.
  Run that document first. The map uses the `species` sheet; the popups use
  `pollinators_long`.
-->

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## What we did this semester

This semester was dedicated to continuing the search through the papers for
latitudes and longitudes in the Americas to complete the workbook, and so keep
advancing the compilation and analysis of the data.

The document is organized into: data preparation, a general **map**, and a set
of **maps by subfamily**. Every map is built from the same data and the same
function, so there is no duplicated code.

The maps below are interactive — pan, zoom, click any point for the species,
locality, traits and named pollinators, and use the checkboxes at the top right
to show or hide each pollinator group. The Shiny drill-down application
(pollinator group → family → genus) lives in `mapa_final.qmd` and needs a live
R session, so it is not part of this shareable file.

***

## Packages

```{r}
#| label: packages
library(readxl)   # read Excel files (.xlsx, .xls)
library(openxlsx) # read the enriched workbook's sheets
library(dplyr)    # data manipulation: filter, mutate, count, the %>% pipe
library(tidyr)    # tidyr::fill() -- carries a value down into blank rows
library(stringr)  # text/pattern helpers, mainly str_detect()
library(leaflet)  # interactive maps
```

***

## Loading and preparing the data

We read the `species` sheet of the enriched workbook and clean the column names.

```{r}
#| label: load-data
# Path to the enriched workbook. This is the ONLY place the file name lives:
# if it is renamed, change just this line.
data_path <- "Pollination_List_pollinators_enriched.xlsx"

# The `species` sheet drives the map; `pollinators_long` drives the drill-down.
data      <- read.xlsx(data_path, sheet = "species")
poll_long <- read.xlsx(data_path, sheet = "pollinators_long")

# Why clean the column names:
# the headers can have inconsistent capitalization and stray trailing spaces.
# We trim and lowercase once here so the rest of the code relies on tidy names.
colnames(data) <- tolower(trimws(colnames(data)))

# A stable row id per record, matching pollinators_long$row (same sheet order).
data$row <- seq_len(nrow(data))
```

We detect the relevant columns **by their name**, not by their position. The
sheet grows and is reordered between semesters; searching by name keeps the
code from breaking if a column moves.

```{r}
#| label: detect-columns
# Helper: given all column names and a search pattern, return the FIRST
# column name that matches, or NA if none matches.
find_col <- function(col_names, pattern) {
  found <- col_names[str_detect(col_names, pattern)]
  if (length(found) == 0) NA_character_ else found[1]
}

col_names <- names(data)
lat_col   <- find_col(col_names, "lat")              # e.g. "latitud"
lon_col   <- find_col(col_names, "lon")              # e.g. "longitud"
# Two pollinator columns, and the order matters -- see the note in the next
# chunk. `poll_groups` is the one derived by the wrangling doc from the
# free-text `pollinators` cell; `pollinator_type` is the older hand-typed column.
grp_col   <- find_col(col_names, "^poll_groups$")    # preferred: the parsed groups
poll_col  <- find_col(col_names, "^pollinator_type$")# fallback: the legacy column
if (is.na(poll_col)) poll_col <- find_col(col_names, "pollinator_type")
loc_col   <- find_col(col_names, "locality|localidad|sitio|lugar")
spec_col  <- find_col(col_names, "^species$|especie|taxon")
# "^genus$" matches the column named EXACTLY "genus". A loose "genus" pattern
# would also catch "genus count", which is a different column.
genus_col <- find_col(col_names, "^genus$")
subf_col  <- find_col(col_names, "^subfamily$")

# Coordinates are essential. If they are missing, stop now with a clear message.
if (is.na(lat_col) || is.na(lon_col)) {
  stop("No latitude/longitude columns were found in the 'species' sheet.")
}

cat("Latitude column:  ", lat_col,   "\n")
cat("Longitude column: ", lon_col,   "\n")
cat("Pollinator group column (preferred):", grp_col,  "\n")
cat("Pollinator type column (fallback):  ", poll_col, "\n")
cat("Genus column:     ", genus_col, "\n")
cat("Subfamily column: ", subf_col,  "\n")
```

We build the map dataset.

```{r}
#| label: prepare-data
# Coordinates arrive as TEXT, sometimes with stray spaces or a comma as the
# decimal separator ("18,234"). This helper turns that text into clean numbers.
to_number <- function(x) {
  x <- gsub("\\s+", "", x)  # remove any spaces
  x <- gsub(",", ".", x)    # comma -> dot, because R needs a dot decimal
  suppressWarnings(as.numeric(x))
}

# Carry `genus`/`subfamily` down into their blank rows (idempotent if already
# filled). MUST run before filtering, so the original row order stays intact.
fill_down <- function(df, col) {
  if (is.na(col)) return(df)
  v <- trimws(as.character(df[[col]]))
  df[[col]] <- ifelse(v == "" | is.na(v), NA_character_, v)
  tidyr::fill(df, all_of(col), .direction = "down")
}

data <- data %>% fill_down(genus_col) %>% fill_down(subf_col)

map_data <- data %>%
  mutate(
    lat = to_number(.data[[lat_col]]),
    lon = to_number(.data[[lon_col]]),

    # Every pollinator group recorded for this record, e.g. "Bees; Diptera".
    #
    # WHY `poll_groups` AND NOT `pollinator_type`:
    # `pollinator_type` is the old hand-typed column and it was only ever filled
    # in for 99 of the 3,161 rows (64 of the 171 georeferenced ones). Reading it
    # alone labelled 107 of the 171 mapped points "Not reported" -- but 65 of
    # those 107 records DO name their pollinators, by species, in the free-text
    # `pollinators` column. `poll_groups` is what the wrangling doc derives by
    # parsing that free text, so it covers 126 of the 171. We read it first and
    # keep `pollinator_type` only as a fallback for the handful of rows the
    # parser could not resolve.
    pollinator_all = {
      g <- if (!is.na(grp_col)) trimws(as.character(.data[[grp_col]])) else NA_character_
      ifelse(is.na(g) | g %in% c("", "NA"), NA_character_, g)
    },

    # The single group used for the point colour and its map layer. A record
    # with several groups is plotted under the FIRST one, so every point belongs
    # to exactly one layer; the popup still lists all of them.
    pollinator = {
      primary <- sub(";.*$", "", pollinator_all)

      # Fallback: the legacy column, standardized to full taxonomic names.
      v <- tolower(trimws(as.character(.data[[poll_col]])))
      recode_poll <- c(bees = "Bees", diptera = "Diptera", wasp = "Wasps",
                       wasps = "Wasps", butterfly = "Lepidoptera",
                       lepid = "Lepidoptera", coleo = "Coleoptera")
      legacy <- unname(recode_poll[v])                                             # known labels
      legacy <- ifelse(is.na(legacy) & !is.na(v) & nzchar(v), str_to_title(v), legacy)

      out <- ifelse(!is.na(primary), primary, legacy)
      # "Not reported" now means what it says: no pollinator recorded anywhere.
      ifelse(is.na(out), "Not reported", out)
    },

    # Full scientific name = genus + epithet (genus was filled down above;
    # the `species` column holds only the epithet).
    sci_name = {
      gen <- if (!is.na(genus_col)) as.character(.data[[genus_col]]) else NA_character_
      epi <- if (!is.na(spec_col))  as.character(.data[[spec_col]])  else NA_character_
      trimws(paste(ifelse(is.na(gen), "", gen), ifelse(is.na(epi), "", epi)))
    },
    locality = if (!is.na(loc_col)) as.character(.data[[loc_col]]) else NA_character_,

    subfamily_label = {
      s <- if (!is.na(subf_col)) as.character(.data[[subf_col]]) else NA_character_
      ifelse(is.na(s) | s == "", "No subfamily", s)
    }
  ) %>%
  filter(!is.na(lat), !is.na(lon))

# Keep only the pollinator rows that belong to a georeferenced (mapped) record,
# so the drill-down menus only offer pollinators that actually appear on the map.
poll_long <- poll_long %>% filter(row %in% map_data$row)

# For each mapped record, an HTML block listing its pollinator SPECIES, grouped
# by family, so the popup shows the named pollinators when available.
poll_popup <- poll_long %>%
  mutate(name = case_when(
    rank == "taxon" ~ paste0("<i>", poll_taxon, "</i>"),
    rank == "genus" ~ paste0("<i>", poll_genus, "</i> sp."),
    TRUE            ~ verbatim)) %>%
  group_by(row, poll_family) %>%
  summarise(line = paste0(ifelse(is.na(poll_family[1]), "", paste0(poll_family[1], ": ")),
                          paste(unique(name), collapse = ", ")), .groups = "drop_last") %>%
  summarise(poll_species = paste(line, collapse = "<br>"), .groups = "drop")

map_data <- map_data %>% left_join(poll_popup, by = "row")

# Build the popup. A trait row is shown only when that trait is recorded.
trait_row <- function(label, value) ifelse(is.na(value), "",
  paste0("<b>", label, ":</b> ", value, "<br>"))

map_data <- map_data %>%
  mutate(popup_info = paste0(
    "<div style='font-size:13px;max-width:330px'>",
    "<div style='font-size:15px;margin-bottom:2px'><b><i>", sci_name, "</i></b></div>",
    "<b>Subfamily:</b> ", subfamily_label, "<br>",
    "<b>Locality:</b> ", ifelse(is.na(locality), "—", locality), "<br>",
    "<b>Coordinates:</b> ", round(lat, 4), " , ", round(lon, 4), "<br>",
    trait_row("Growth form",     growth_form),
    trait_row("Climate",         climate),
    trait_row("Compatibility",   compatibility),
    trait_row("Breeding system", breeding_system),
    trait_row("Reward",          reward),
    trait_row("Deception",       deception),
    ifelse(is.na(mycoheterotrophic), "", paste0("<b>", mycoheterotrophic, "</b><br>")),
    # Show the full group list only when there is more than one, since a single
    # group is already obvious from the point's colour.
    ifelse(is.na(pollinator_all) | !grepl(";", pollinator_all), "",
           paste0("<b>Pollinator groups:</b> ", gsub("; *", ", ", pollinator_all), "<br>")),
    "<b>Pollinators:</b> ",
    ifelse(is.na(poll_species) | poll_species == "",
           paste0(pollinator, " <span style='color:#888'>(no species recorded)</span>"),
           paste0("<br>", poll_species)),
    "</div>"
  ))
```

### Quick summary

```{r}
#| label: summary
cat("Records with valid coordinates:", nrow(map_data), "\n")
cat("Distinct pollinator types:     ", dplyr::n_distinct(map_data$pollinator), "\n")
```

```{r}
#| label: pollinator-table
map_data %>%
  count(pollinator, name = "records") %>%
  arrange(desc(records)) %>%
  knitr::kable(caption = "Records by pollinator type")
```

***

## How the maps are built

We define a single colour palette and a single function, `build_map()`, used by
every map below, so the map is described in just one place.

```{r}
#| label: map-function
# Fixed colour per pollinator group (colourblind-friendly Okabe-Ito), the SAME
# palette as pollinator_map.qmd so both maps look alike. Each group's colour is
# used for its map points AND for the coloured dot on its layer-control
# checkbox, so the checkboxes double as the legend (no separate legend box).
group_colors <- c(Bees = "#E69F00", Diptera = "#56B4E9", Wasps = "#D55E00",
                  Coleoptera = "#009E73", Lepidoptera = "#CC79A7", Aves = "#0072B2",
                  Hemiptera = "#F0E442", "Hymenoptera (unassigned)" = "#8C6D31",
                  "Not reported" = "#BBBBBB")
color_of <- function(g) unname(ifelse(g %in% names(group_colors), group_colors[g], "#999999"))

# build_map(): takes a prepared data frame and returns a Leaflet map. Defining
# it once means the overall map and the per-subfamily maps cannot drift apart.
build_map <- function(df) {
  groups <- sort(unique(df$pollinator))

  # Each layer's label is a coloured dot (same colour as its points) followed
  # by the group name; used as BOTH the marker group and the checkbox label.
  swatch <- function(g) sprintf(
    "<span style='display:inline-block;width:11px;height:11px;border-radius:50%%;background:%s;border:1px solid #444;margin-right:5px;vertical-align:middle'></span>%s",
    color_of(g), g)
  labels <- setNames(vapply(groups, swatch, character(1)), groups)

  m <- leaflet(df) %>%
    addTiles() %>%                                  # standard street base map
    addProviderTiles(providers$Esri.WorldImagery)   # satellite imagery on top

  # One marker layer per pollinator group, each in its own colour. The layer
  # name (`group=`) is the coloured HTML label, which the layer control renders.
  for (pol in groups) {
    m <- m %>%
      addCircleMarkers(
        data = df %>% filter(pollinator == pol),
        lng = ~lon, lat = ~lat,
        popup = ~popup_info,
        radius = 5, stroke = TRUE, weight = 1, fillOpacity = 0.7,
        color = color_of(pol),
        group = labels[[pol]]
      )
  }

  # The control box (top-right): a coloured, toggleable checkbox per group.
  m %>%
    addLayersControl(
      overlayGroups = unname(labels[groups]),
      position = "topright",
      options = layersControlOptions(collapsed = FALSE)
    )
}
```

## Map of all records

Each point is a locality with its reported pollinator type.

```{r}
#| label: overall-map
build_map(map_data)
```

## Maps by subfamily

One map per subfamily that has georeferenced records. The colours still
represent the pollinator type, just like the general map, so the maps are
comparable with one another.

```{r}
#| label: subfamily-maps
htmltools::tagList(
  lapply(sort(unique(map_data$subfamily_label)), function(sf) {
    htmltools::tagList(
      htmltools::h4(sf),
      build_map(filter(map_data, subfamily_label == sf))
    )
  })
)
```

***
