---
title: "Orchid Pollinators — Global Distribution"
author: "Raymond L. Tremblay, Natalia Palou, Caleb Pacheco and Naan Lamoso"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-location: left
    theme: flatly
    df-print: paged
    code-fold: true          # code chunks start collapsed...
    code-summary: "Show code" # ...with this label to expand them
    code-tools: true         # global button to show/hide all code
    embed-resources: true    # inline every JS/CSS asset -> ONE portable .html
lang: en
---

<!--
  STATIC, SELF-CONTAINED version of `mapa_final.qmd`, for sharing by email.

  Differences from `mapa_final.qmd`:
    * no `server: shiny` -> renders with a plain `quarto render`, no live R
      session needed to view it;
    * `embed-resources: true` -> every JS/CSS asset is inlined, so the output
      is ONE .html file with no `_files/` folder beside it;
    * the Shiny drill-down app at the end is removed (it cannot exist in a
      static file). The static map and the per-subfamily maps are unchanged and
      fully interactive: pan, zoom, click a point for its popup, and toggle
      pollinator groups with the checkboxes.

  Render with:  quarto render mapa_final_static.qmd
  Then email `mapa_final_static.html` on its own. Do NOT send the workbook --
  the data is baked into the file.

  NOTE: the base map tiles (OpenStreetMap + Esri satellite) stream from the web,
  so the recipient needs an internet connection when they open it. Everything
  else -- points, popups, data -- is inside the file.

  Data source: the ENRICHED workbook produced by `pollinators_wrangling.qmd`.
  Run that document first. The map uses the `species` sheet; the popups use
  `pollinators_long`.
-->

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

Orchidaceae is among the largest families of flowering plants, and its
pollination systems are correspondingly diverse. Records of who pollinates what,
however, are scattered across a century of primary literature and are rarely
tied to explicit coordinates, which makes the geography of orchid pollination
difficult to examine directly.

This document maps the georeferenced portion of a global compilation of orchid
pollination records. The underlying database was assembled by Ackerman *et al.*
(2023) and covers roughly 3,100 species; the work presented here adds
coordinates to those records, species by species, from the localities reported
in the source publications.

Georeferencing is in progress and proceeds by region, the Americas first and
Asia to follow. The maps below therefore show the records resolved to date
rather than the full database, and coverage is uneven between subfamilies.

Each point is interactive: pan and zoom, click a marker for the species,
locality, traits and named pollinators, and use the checkboxes at the upper
right to show or hide individual pollinator groups.

### Data source

Pollination records are drawn from the global orchid pollination database of
Ackerman JD, Phillips RD, Tremblay RL, Karremans A, Reiter N, Peter CI, Bogarín
D, Pérez-Escobar OA & Liu H (2023), *Beyond the various contrivances by which
orchids are pollinated: global patterns in orchid pollination biology*,
**Botanical Journal of the Linnean Society** 202: 295–324, archived at
<https://doi.org/10.5281/zenodo.7263689> under CC BY 4.0.

***

## Packages

```{r}
#| label: packages
library(readxl)   # read Excel files (.xlsx, .xls)
library(openxlsx) # read the enriched workbook's sheets
library(dplyr)    # data manipulation: filter, mutate, count, the %>% pipe
library(tidyr)    # tidyr::fill() -- carries a value down into blank rows
library(stringr)  # text/pattern helpers, mainly str_detect()
library(leaflet)  # interactive maps
```

***

## Loading and preparing the data

The `species` sheet of the enriched workbook is read and its column names are
normalised.

```{r}
#| label: load-data
# Path to the enriched workbook. This is the ONLY place the file name lives:
# if it is renamed, change just this line.
data_path <- "Pollination_List_pollinators_enriched.xlsx"

# The `species` sheet drives the map; `pollinators_long` drives the drill-down.
data      <- read.xlsx(data_path, sheet = "species")
poll_long <- read.xlsx(data_path, sheet = "pollinators_long")

# Why clean the column names:
# the headers can have inconsistent capitalization and stray trailing spaces.
# We trim and lowercase once here so the rest of the code relies on tidy names.
colnames(data) <- tolower(trimws(colnames(data)))

# A stable row id per record, matching pollinators_long$row (same sheet order).
data$row <- seq_len(nrow(data))
```

Columns are identified **by name rather than by position**. The workbook grows
and is periodically reordered, so matching on name keeps the code stable when a
column moves.

```{r}
#| label: detect-columns
# Helper: given all column names and a search pattern, return the FIRST
# column name that matches, or NA if none matches.
find_col <- function(col_names, pattern) {
  found <- col_names[str_detect(col_names, pattern)]
  if (length(found) == 0) NA_character_ else found[1]
}

col_names <- names(data)
lat_col   <- find_col(col_names, "lat")              # e.g. "latitud"
lon_col   <- find_col(col_names, "lon")              # e.g. "longitud"
# Two pollinator columns, and the order matters -- see the note in the next
# chunk. `poll_groups` is the one derived by the wrangling doc from the
# free-text `pollinators` cell; `pollinator_type` is the older hand-typed column.
grp_col   <- find_col(col_names, "^poll_groups$")    # preferred: the parsed groups
poll_col  <- find_col(col_names, "^pollinator_type$")# fallback: the legacy column
if (is.na(poll_col)) poll_col <- find_col(col_names, "pollinator_type")
loc_col   <- find_col(col_names, "locality|localidad|sitio|lugar")
spec_col  <- find_col(col_names, "^species$|especie|taxon")
# "^genus$" matches the column named EXACTLY "genus". A loose "genus" pattern
# would also catch "genus count", which is a different column.
genus_col <- find_col(col_names, "^genus$")
subf_col  <- find_col(col_names, "^subfamily$")

# Coordinates are essential. If they are missing, stop now with a clear message.
if (is.na(lat_col) || is.na(lon_col)) {
  stop("No latitude/longitude columns were found in the 'species' sheet.")
}

cat("Latitude column:  ", lat_col,   "\n")
cat("Longitude column: ", lon_col,   "\n")
cat("Pollinator group column (preferred):", grp_col,  "\n")
cat("Pollinator type column (fallback):  ", poll_col, "\n")
cat("Genus column:     ", genus_col, "\n")
cat("Subfamily column: ", subf_col,  "\n")
```

```{r}
#| label: prepare-helpers
# Coordinates arrive as TEXT, sometimes with stray spaces or a comma as the
# decimal separator ("18,234"). This helper turns that text into clean numbers.
to_number <- function(x) {
  x <- gsub("\\s+", "", x)  # remove any spaces
  x <- gsub(",", ".", x)    # comma -> dot, because R needs a dot decimal
  suppressWarnings(as.numeric(x))
}

# Carry `genus`/`subfamily` down into their blank rows (idempotent if already
# filled). MUST run before filtering, so the original row order stays intact.
fill_down <- function(df, col) {
  if (is.na(col)) return(df)
  v <- trimws(as.character(df[[col]]))
  df[[col]] <- ifelse(v == "" | is.na(v), NA_character_, v)
  tidyr::fill(df, all_of(col), .direction = "down")
}

data <- data %>% fill_down(genus_col) %>% fill_down(subf_col)
```

The map dataset is then assembled: numeric coordinates, the pollinator group
that determines each point's colour, and the labels used in the popups.

Two columns record the pollinator, and the choice between them matters.
`pollinator_type` is a legacy field populated for only 99 of the 3,161 records
(64 of the georeferenced ones). Relying on it alone leaves 107 of the mapped
points unattributed, although 65 of those records name their pollinators to
species in the free-text `pollinators` field. `poll_groups`, derived from that
free text during data preparation, resolves 126 of the georeferenced records and
is therefore read first, with `pollinator_type` retained as a fallback.

```{r}
#| label: prepare-map-data
map_data <- data %>%
  mutate(
    lat = to_number(.data[[lat_col]]),
    lon = to_number(.data[[lon_col]]),

    # Every group recorded for this record, e.g. "Bees; Diptera". See above for
    # why this reads `poll_groups` first and `pollinator_type` only as a fallback.
    pollinator_all = {
      g <- if (!is.na(grp_col)) trimws(as.character(.data[[grp_col]])) else NA_character_
      ifelse(is.na(g) | g %in% c("", "NA"), NA_character_, g)
    },

    # The single group used for the point colour and its map layer. A record
    # with several groups is plotted under the FIRST one, so every point belongs
    # to exactly one layer; the popup still lists all of them.
    pollinator = {
      primary <- sub(";.*$", "", pollinator_all)

      # Fallback: the legacy column, standardized to full taxonomic names.
      v <- tolower(trimws(as.character(.data[[poll_col]])))
      recode_poll <- c(bees = "Bees", diptera = "Diptera", wasp = "Wasps",
                       wasps = "Wasps", butterfly = "Lepidoptera",
                       lepid = "Lepidoptera", coleo = "Coleoptera")
      legacy <- unname(recode_poll[v])                                             # known labels
      legacy <- ifelse(is.na(legacy) & !is.na(v) & nzchar(v), str_to_title(v), legacy)

      out <- ifelse(!is.na(primary), primary, legacy)
      # "Not reported" now means what it says: no pollinator recorded anywhere.
      ifelse(is.na(out), "Not reported", out)
    },

    # Full scientific name = genus + epithet (genus was filled down above;
    # the `species` column holds only the epithet).
    sci_name = {
      gen <- if (!is.na(genus_col)) as.character(.data[[genus_col]]) else NA_character_
      epi <- if (!is.na(spec_col))  as.character(.data[[spec_col]])  else NA_character_
      trimws(paste(ifelse(is.na(gen), "", gen), ifelse(is.na(epi), "", epi)))
    },
    locality = if (!is.na(loc_col)) as.character(.data[[loc_col]]) else NA_character_,

    subfamily_label = {
      s <- if (!is.na(subf_col)) as.character(.data[[subf_col]]) else NA_character_
      ifelse(is.na(s) | s == "", "No subfamily", s)
    }
  ) %>%
  filter(!is.na(lat), !is.na(lon))

```

The named pollinator taxa for each record are collapsed into a single block per
record, grouped by family.

```{r}
#| label: prepare-pollinator-names
# Keep only the pollinator rows that belong to a georeferenced (mapped) record,
# so the drill-down menus only offer pollinators that actually appear on the map.
poll_long <- poll_long %>% filter(row %in% map_data$row)

# For each mapped record, an HTML block listing its pollinator SPECIES, grouped
# by family, so the popup shows the named pollinators when available.
poll_popup <- poll_long %>%
  mutate(name = case_when(
    rank == "taxon" ~ paste0("<i>", poll_taxon, "</i>"),
    rank == "genus" ~ paste0("<i>", poll_genus, "</i> sp."),
    TRUE            ~ verbatim)) %>%
  group_by(row, poll_family) %>%
  summarise(line = paste0(ifelse(is.na(poll_family[1]), "", paste0(poll_family[1], ": ")),
                          paste(unique(name), collapse = ", ")), .groups = "drop_last") %>%
  summarise(poll_species = paste(line, collapse = "<br>"), .groups = "drop")

map_data <- map_data %>% left_join(poll_popup, by = "row")

```

Finally, the popup content itself.

```{r}
#| label: prepare-popup
# Build the popup. A trait row is shown only when that trait is recorded.
trait_row <- function(label, value) ifelse(is.na(value), "",
  paste0("<b>", label, ":</b> ", value, "<br>"))

map_data <- map_data %>%
  mutate(popup_info = paste0(
    "<div style='font-size:13px;max-width:330px'>",
    "<div style='font-size:15px;margin-bottom:2px'><b><i>", sci_name, "</i></b></div>",
    "<b>Subfamily:</b> ", subfamily_label, "<br>",
    "<b>Locality:</b> ", ifelse(is.na(locality), "—", locality), "<br>",
    "<b>Coordinates:</b> ", round(lat, 4), " , ", round(lon, 4), "<br>",
    trait_row("Growth form",     growth_form),
    trait_row("Climate",         climate),
    trait_row("Compatibility",   compatibility),
    trait_row("Breeding system", breeding_system),
    trait_row("Reward",          reward),
    trait_row("Deception",       deception),
    ifelse(is.na(mycoheterotrophic), "", paste0("<b>", mycoheterotrophic, "</b><br>")),
    # Show the full group list only when there is more than one, since a single
    # group is already obvious from the point's colour.
    ifelse(is.na(pollinator_all) | !grepl(";", pollinator_all), "",
           paste0("<b>Pollinator groups:</b> ", gsub("; *", ", ", pollinator_all), "<br>")),
    "<b>Pollinators:</b> ",
    ifelse(is.na(poll_species) | poll_species == "",
           paste0(pollinator, " <span style='color:#888'>(no species recorded)</span>"),
           paste0("<br>", poll_species)),
    "</div>"
  ))
```

### Summary

```{r}
#| label: summary
cat("Records with valid coordinates:", nrow(map_data), "\n")
cat("Distinct pollinator types:     ", dplyr::n_distinct(map_data$pollinator), "\n")
```

```{r}
#| label: pollinator-table
map_data %>%
  count(pollinator, name = "records") %>%
  arrange(desc(records)) %>%
  knitr::kable(caption = "Records by pollinator type")
```

***

## Mapping

A single colour palette and a single function, `build_map()`, serve every map in
this document, so the cartography is defined in one place.

```{r}
#| label: map-function
# Fixed colour per pollinator group (colourblind-friendly Okabe-Ito), the SAME
# palette as pollinator_map.qmd so both maps look alike. Each group's colour is
# used for its map points AND for the coloured dot on its layer-control
# checkbox, so the checkboxes double as the legend (no separate legend box).
group_colors <- c(Bees = "#E69F00", Diptera = "#56B4E9", Wasps = "#D55E00",
                  Coleoptera = "#009E73", Lepidoptera = "#CC79A7", Aves = "#0072B2",
                  Hemiptera = "#F0E442", "Hymenoptera (unassigned)" = "#8C6D31",
                  "Not reported" = "#BBBBBB")
color_of <- function(g) unname(ifelse(g %in% names(group_colors), group_colors[g], "#999999"))

```

`build_map()` converts a prepared data frame into a Leaflet map.

```{r}
#| label: build-map-function
# build_map(): takes a prepared data frame and returns a Leaflet map. Defining
# it once means the overall map and the per-subfamily maps cannot drift apart.
build_map <- function(df) {
  groups <- sort(unique(df$pollinator))

  # Each layer's label is a coloured dot (same colour as its points) followed
  # by the group name; used as BOTH the marker group and the checkbox label.
  swatch <- function(g) sprintf(
    "<span style='display:inline-block;width:11px;height:11px;border-radius:50%%;background:%s;border:1px solid #444;margin-right:5px;vertical-align:middle'></span>%s",
    color_of(g), g)
  labels <- setNames(vapply(groups, swatch, character(1)), groups)

  m <- leaflet(df) %>%
    addTiles() %>%                                  # standard street base map
    addProviderTiles(providers$Esri.WorldImagery)   # satellite imagery on top

  # One marker layer per pollinator group, each in its own colour. The layer
  # name (`group=`) is the coloured HTML label, which the layer control renders.
  for (pol in groups) {
    m <- m %>%
      addCircleMarkers(
        data = df %>% filter(pollinator == pol),
        lng = ~lon, lat = ~lat,
        popup = ~popup_info,
        radius = 5, stroke = TRUE, weight = 1, fillOpacity = 0.7,
        color = color_of(pol),
        group = labels[[pol]]
      )
  }

  # The control box (top-right): a coloured, toggleable checkbox per group.
  m %>%
    addLayersControl(
      overlayGroups = unname(labels[groups]),
      position = "topright",
      options = layersControlOptions(collapsed = FALSE)
    )
}
```

## Distribution of all georeferenced records

Each point marks a locality at which pollination was observed, coloured by the
pollinator group recorded there.

```{r}
#| label: overall-map
build_map(map_data)
```

## Maps by subfamily

One map per subfamily with georeferenced records. Colours follow the same
pollinator scheme as above, so the panels are directly comparable.

```{r}
#| label: subfamily-maps
htmltools::tagList(
  lapply(sort(unique(map_data$subfamily_label)), function(sf) {
    htmltools::tagList(
      htmltools::h4(sf),
      build_map(filter(map_data, subfamily_label == sf))
    )
  })
)
```

***
