---
title: "Cleaning and structuring the *pollinators* column"
subtitle: "Global Orchid Pollinators — preparing the data for search and Leaflet"
author: "Natalia Palou and Caleb Pacheco"
date: "2026-08-13"
format:
  html:
    toc: true
    toc-float: true
    theme: flatly
    df-print: paged
    code-fold: show
lang: en
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Goal

The **`pollinators`** column (column AF in the Excel) stores the pollinators as
free text, following a fairly consistent grammar:

```
ORDER:  Genus species, G. species2 (Family); Genus2 species (Family2).  ORDER2:  ...
```

That is: one or more **orders** in UPPERCASE followed by `:`; within each order,
a comma-separated list of taxa (often abbreviated after the first mention,
`Trigona melanocephala, T. minangkabau`), grouped into `(Family)`; families are
separated by `;` and orders by `.`.

The challenge we wanted to solve is twofold and seemingly contradictory: when a
record has **more than one pollinator** we need to be able to **separate** them
(to search by a specific genus or family) and at the same time **amalgamate**
them (so each point on the map can show a summary). The solution is to do a
**single parse** and derive **two views** from it:

1. A **long table** (`pollinators_long`): one row per *orchid × pollinator
   taxon*. This is the *separated* view, searchable at any rank (order, family,
   genus, species).
2. A set of **wide columns** (`poll_*`) joined back onto each orchid row. This
   is the *amalgamated* view, and it feeds the map popup and filters. They are
   generated **from** the long table, so the two never fall out of sync.

The original `pollinators` column is **left untouched**: it is the raw source.
All cleaning (including order and family typos) lives here, in the code — the
same philosophy `mapa_final.rmd` already uses with the `wasp → wasps` recode.

***

## Packages

```{r libraries}
library(readxl)    # read the Excel
library(dplyr)     # data manipulation
library(tidyr)     # fill(), unnest()
library(stringr)   # regular expressions
library(purrr)     # map()
library(openxlsx)  # write the enriched workbook (multiple sheets)
```

## Loading the data

We read the `species` sheet of the merged workbook and clean the column names,
just like the main analysis.

```{r load}
# The only line with the file name: if it is renamed, change it here.
ruta_datos <- "Pollination_List_Thru_1 2024_merged26_08_13.xlsx"

data <- read_excel(ruta_datos, sheet = "species")
colnames(data) <- tolower(trimws(colnames(data)))

# Clean data-entry errors in `subfamily` BEFORE carrying it down. Orchidaceae
# has exactly five subfamilies; everything else here is an error:
#   "2"           -> Epidendroideae  (those rows are Bulbophyllum)
#   "Orchidaceae" -> Orchidoideae    (those rows are Sirindhornia)
#   "5" / "subfamily" -> NA          (junk header rows, genus == "genus")
subf_fix <- c("2" = "Epidendroideae", "Orchidaceae" = "Orchidoideae",
              "5" = NA_character_, "subfamily" = NA_character_)
sv <- trimws(as.character(data$subfamily))
data$subfamily <- ifelse(sv %in% names(subf_fix), unname(subf_fix[sv]), sv)

# Carry `genus` and `subfamily` down (they are written only once). This must
# run before any row filtering so the original row order stays intact.
fill_down <- function(df, col) {
  v <- trimws(as.character(df[[col]]))
  df[[col]] <- ifelse(v == "" | is.na(v), NA_character_, v)
  tidyr::fill(df, all_of(col), .direction = "down")
}
data <- data %>% fill_down("genus") %>% fill_down("subfamily")

# Full scientific name (genus + epithet) and a row id to rejoin on later.
data$orchid <- str_squish(paste(ifelse(is.na(data$genus), "", data$genus),
                                ifelse(is.na(data$species), "", data$species)))
data$.rowid <- seq_len(nrow(data))
```

## Species traits

The workbook also carries many one-hot trait columns (a `1` marks the trait,
blank means absent): growth form, climate zone, breeding system, and floral
rewards / deception strategy. We turn those scattered flags into a few
readable, searchable columns. Absent traits stay `NA` rather than "0", so
"unknown" and "recorded absent" are not confused.

```{r traits}
# TRUE only when a flag cell literally holds 1 / TRUE.
is_present <- function(x) tolower(trimws(as.character(x))) %in% c("1", "true")
flag <- function(name) if (name %in% names(data)) is_present(data[[name]]) else rep(FALSE, nrow(data))

# Combine several flag columns into one "A; B" label column (NA if none set).
combine_flags <- function(specs) {           # specs: c(Label = "column_name", ...)
  out <- rep("", nrow(data))
  for (k in seq_along(specs)) {
    on <- flag(unname(specs[k]))
    out <- ifelse(on, ifelse(out == "", names(specs)[k], paste(out, names(specs)[k], sep = "; ")), out)
  }
  ifelse(out == "", NA_character_, out)
}

data$growth_form   <- combine_flags(c(Epiphytic = "epi", Terrestrial = "terr",
                                      Rupicolous = "rupic", Climbing = "vine"))
data$climate       <- combine_flags(c(Tropical = "trop", Subtropical = "subtr",
                                      Temperate = "temp"))
data$compatibility <- combine_flags(c(`Self-incompatible` = "si", `Self-compatible` = "sc"))
data$breeding_system <- combine_flags(c(Chasmogamous = "chasmogamous",
                                       `Mixed mating` = "mixed mating",
                                       `Autonomous selfing/agamospermy` = "autonomous selfing/agagamospermy"))
data$reward        <- combine_flags(c(Nectar = "nectar", Pollen = "pollen",
                                     Fragrance = "fragr", `Oils/lipids` = "lipids",
                                     Pseudopollen = "trichomes/pseudo-pollen",
                                     `Sleeping site` = "sleep sites", Resins = "polyphenols resins"))
data$deception     <- combine_flags(c(`Food deception` = "food dec",
                                      `Sexual deception` = "sex dec",
                                      `Brood-site deception` = "brood site dec"))
data$mycoheterotrophic <- ifelse(flag("achlorophyllous"), "Mycoheterotrophic", NA_character_)

# How many records carry each derived trait (a quick fill check).
sapply(c("growth_form","climate","compatibility","breeding_system",
         "reward","deception","mycoheterotrophic"),
       function(c) sum(!is.na(data[[c]])))
```

## Recode maps (typos)

Orders have ~10 real values but ~20 different spellings; families have similar
typos. We fix them with two small dictionaries that are easy to extend as new
variants appear.

```{r recodes}
order_recode <- c(
  DIPERA = "DIPTERA", HYEMNOPTERA = "HYMENOPTERA", HHYMENOPTERA = "HYMENOPTERA",
  HYMENOTERA = "HYMENOPTERA", HYMENOTPERA = "HYMENOPTERA",
  LEPIODOPTERA = "LEPIDOPTERA", LEPTIDOPTERA = "LEPIDOPTERA", HEMPTERIA = "HEMIPTERA")

family_recode <- c(
  Andrinidae = "Andrenidae", Aipdae = "Apidae", Mellitidae = "Melittidae",
  Anthomyidae = "Anthomyiidae", Choropidae = "Chloropidae", Hesperidae = "Hesperiidae",
  Tephrididae = "Tephritidae", Empidae = "Empididae", Papillionidae = "Papilionidae",
  Papilioidae = "Papilionidae", Cramabidae = "Crambidae", Nymphaidae = "Nymphalidae",
  Bombylidae = "Bombyliidae", Calliphridae = "Calliphoridae", Syriphidae = "Syrphidae",
  Thynnindae = "Thynnidae", Thinnidae = "Thynnidae", Thyniidae = "Thynnidae",
  Thiniidae = "Thynnidae", Pompiliidae = "Pompilidae", Pompylidae = "Pompilidae",
  Scatophagidae = "Scathophagidae", Chrysidae = "Chrysididae", Scolidae = "Scoliidae",
  Cullicidae = "Culicidae", Mycetophiilidae = "Mycetophilidae", Saturnidae = "Saturniidae",
  Tabinidae = "Tabanidae", Tenthredionidae = "Tenthredinidae", Erabidae = "Erebidae",
  Zosteropsidae = "Zosteropidae", Ullidiidae = "Ulidiidae", Apidea = "Apidae")

fix_order  <- function(o) { o <- toupper(o); ifelse(o %in% names(order_recode),
                                                    unname(order_recode[o]), o) }
fix_family <- function(x) ifelse(!is.na(x) & x %in% names(family_recode),
                                 unname(family_recode[x]), x)

# Coarse group, using full taxonomic names (no contractions). Whole orders
# keep the order name; HYMENOPTERA is split into Bees vs. Wasps, because that
# distinction matters for pollination ecology.
bee_fams <- c("Apidae","Halictidae","Andrenidae","Megachilidae","Colletidae",
              "Melittidae","Stenotritidae")

# Family is the primary signal, but many source papers name the bee without its
# family ("HYMENOPTERA: Bombus cf. albopleuralis"), or misspell it. Falling
# straight through to "Wasps" in that case mislabels obvious bees, so we check
# the GENUS as a second signal before giving up. Extend this list as needed.
bee_genera <- c("Apis","Bombus","Xylocopa","Euglossa","Eulaema","Eufriesea",
                "Exaerete","Melipona","Trigona","Tetragonisca","Nannotrigona",
                "Scaptotrigona","Centris","Ceratina","Tetralonia","Tetraloniella",
                "Eucera","Anthophora","Andrena","Nomada","Lasioglossum","Halictus",
                "Augochlora","Augochlorella","Augochloropsis","Megachile","Osmia",
                "Colletes","Hylaeus","Melissodes","Peponapis","Xylocopinae")

group_from <- function(order, family, genus = NA_character_) {
  order <- toupper(order); if (is.na(order)) return(NA_character_)
  switch(order,
    DIPTERA = "Diptera", COLEOPTERA = "Coleoptera", LEPIDOPTERA = "Lepidoptera",
    HEMIPTERA = "Hemiptera", THYSANOPTERA = "Thysanoptera", ORTHOPTERA = "Orthoptera",
    ARANEAE = "Arachnida", ARACHNIDA = "Arachnida", AVES = "Aves",
    HYMENOPTERA =
      if (!is.na(family) && family %in% bee_fams) "Bees"
      else if (!is.na(genus)  && genus  %in% bee_genera) "Bees"
      # Neither family nor genus identifies it. "Wasps" would be a guess, so say
      # so plainly instead -- these rows are the ones worth checking by hand.
      else if (is.na(family) && is.na(genus)) "Hymenoptera (unassigned)"
      else "Wasps",
    str_to_title(tolower(order)))   # any other order -> proper-case name
}
```

## Parsing function

`parse_poll()` takes **one** `pollinators` cell and returns a table with one
row per taxon: order, family, genus, epithet, full binomial, the original text
(`verbatim`), the rank reached (`taxon` / `genus` / `informal`) and the coarse
group.

```{r parser}
# Common English words that are NOT scientific names (they prevent false
# genera/species such as "Stingless bees" or "fruit flies").
COMMON <- c("stingless","fruit","flies","bees","wasps","moths","beetles",
            "birds","ants","hawk","small","large")

parse_poll <- function(s) {
  if (is.na(s) || !nzchar(trimws(s))) return(tibble())
  s <- str_squish(s)
  # ORDER markers: an UPPERCASE word (>=4 letters) followed by ':'
  m <- str_locate_all(s, "[A-Z]{4,}\\s*:")[[1]]
  if (nrow(m) == 0)
    return(tibble(order = NA_character_, family = NA_character_, genus = NA_character_,
                  epithet = NA_character_, taxon = NA_character_, verbatim = s,
                  rank = "informal"))
  starts <- m[,1]; endc <- m[,2]
  onames <- trimws(str_replace(str_sub(s, starts, endc), ":$", ""))
  bend <- c(starts[-1] - 1, nchar(s)); last_genus <- NA_character_; out <- list()

  for (i in seq_along(onames)) {
    order <- fix_order(onames[i])
    body  <- str_squish(str_sub(s, endc[i] + 1, bend[i]))
    if (!nzchar(body)) next
    # Each segment = one family. Cut after a ')' followed by ';' or '.', and
    # also on lone ';' (taxa with no family in parentheses).
    segs <- unlist(str_split(str_split(body, "(?<=\\))\\s*[;.]+\\s*")[[1]], ";"))
    for (seg in segs) {
      seg <- str_squish(str_replace(seg, "[\\.\\s]+$", ""))
      if (!nzchar(seg)) next
      fam <- NA_character_
      par <- str_match(seg, "\\(([^)]*)\\)\\s*$")          # last parenthesis
      if (!is.na(par[1,2])) fam <- fix_family(str_extract(par[1,2], "[A-Z][a-z]+dae"))
      taxa <- str_squish(str_replace(seg, "\\([^)]*\\)\\s*$", ""))  # drop the family

      for (tok in str_split(taxa, ",")[[1]]) {
        tok <- str_squish(tok); if (!nzchar(tok)) next
        genus <- NA_character_; epithet <- NA_character_; rank <- "taxon"
        if (str_detect(tok, "^[A-Z]\\.\\s*")) {            # abbreviated genus "B. ..."
          rest  <- str_squish(str_replace(tok, "^[A-Z]\\.\\s*", ""))
          genus <- last_genus                              # expand with the last full genus
          epithet <- str_extract(rest, "^[a-z][a-z-]+")
        } else if (str_detect(tok, "^[A-Z][a-z]+")) {      # full genus
          genus <- str_extract(tok, "^[A-Z][a-z]+"); last_genus <- genus
          rest  <- str_squish(str_replace(tok, "^[A-Z][a-z]+\\s*", ""))
          epithet <- str_extract(rest, "^[a-z][a-z-]+")
        } else rank <- "informal"
        # sp./spp./aff./cf. -> genus level
        if (!is.na(epithet) && epithet %in% c("sp","spp","aff","cf","near")) epithet <- NA_character_
        if (!is.na(epithet) && epithet %in% COMMON) { epithet <- NA_character_; rank <- "informal" }
        if (!is.na(genus)   && tolower(genus) %in% COMMON) { genus <- NA_character_; rank <- "informal" }
        if (is.na(genus) && rank == "taxon") rank <- "informal"
        if (rank == "taxon" && is.na(epithet)) rank <- "genus"
        taxon <- if (rank == "taxon") paste(genus, epithet) else if (rank == "genus") genus else NA_character_
        out[[length(out) + 1]] <- tibble(order = order, family = fam, genus = genus,
                                         epithet = epithet, taxon = taxon,
                                         verbatim = tok, rank = rank)
      }
    }
  }
  if (length(out) == 0)
    return(tibble(order = NA, family = NA, genus = NA, epithet = NA, taxon = NA,
                  verbatim = s, rank = "informal"))
  bind_rows(out) %>% mutate(group = pmap_chr(list(order, family, genus), group_from))
}
```

## Long table (the *separated* view)

```{r long}
nonempty <- data %>% filter(!is.na(pollinators) & nzchar(trimws(pollinators)))

long <- nonempty %>%
  mutate(pp = map(pollinators, parse_poll)) %>%
  transmute(row = .rowid, orchid, subfamily, pollinator_type, pollinators, pp) %>%
  unnest(pp) %>%
  transmute(row, orchid, subfamily, pollinator_type,
            poll_order = order, poll_family = family, poll_genus = genus,
            poll_epithet = epithet, poll_taxon = taxon, poll_group = group,
            rank, verbatim)

head(long, 12)
```

## Wide columns (the *amalgamated* view)

These are built by grouping the long table by orchid record, so they are a
faithful summary of the separated data.

```{r wide}
j <- function(x) paste(sort(unique(x[!is.na(x) & nzchar(x)])), collapse = "; ")

wide <- long %>%
  group_by(row) %>%
  summarise(
    poll_orders   = j(poll_order),
    poll_families = j(poll_family),
    poll_genera   = j(poll_genus),
    poll_taxa     = j(poll_taxon),
    poll_groups   = j(poll_group),
    n_orders   = n_distinct(poll_order[!is.na(poll_order)]),
    n_families = n_distinct(poll_family[!is.na(poll_family)]),
    n_genera   = n_distinct(poll_genus[!is.na(poll_genus)]),
    n_taxa     = n_distinct(poll_taxon[rank == "taxon"]),
    grp_bees    = any(poll_group == "Bees"),
    grp_diptera = any(poll_group == "Diptera"),
    grp_wasps   = any(poll_group == "Wasps"),
    grp_coleo   = any(poll_group == "Coleoptera"),
    grp_lepid   = any(poll_group == "Lepidoptera"),
    .groups = "drop")

# Enriched `species` sheet = original columns + wide columns.
species_out <- data %>%
  left_join(wide, by = c(".rowid" = "row")) %>%
  select(-.rowid, -orchid)

head(wide, 8)
```

## Quality control

```{r qc}
cat("Long rows (taxa):", nrow(long), "\n")
cat("Orchid records with pollinators:", nrow(wide), "\n")
cat("Orders:", n_distinct(long$poll_order, na.rm = TRUE),
    "| Families:", n_distinct(long$poll_family, na.rm = TRUE),
    "| Genera:", n_distinct(long$poll_genus, na.rm = TRUE),
    "| Full binomials:", n_distinct(long$poll_taxon[long$rank == "taxon"]), "\n\n")

long %>% count(rank, name = "rows") %>% knitr::kable(caption = "Rows by rank reached")
long %>% count(poll_group, name = "rows") %>% arrange(desc(rows)) %>%
  knitr::kable(caption = "Rows by coarse group")
```

We compare our counts with the columns that were already tallied by hand in the
Excel (`orders`, `families`, `genera`): a high correlation confirms the parse
reproduces the original criterion.

```{r qc-cross}
chk <- long %>% group_by(row) %>%
  summarise(p_ord = n_distinct(poll_order, na.rm = TRUE),
            p_fam = n_distinct(poll_family, na.rm = TRUE),
            p_gen = n_distinct(poll_genus, na.rm = TRUE), .groups = "drop") %>%
  left_join(nonempty %>% transmute(row = .rowid,
              m_ord = suppressWarnings(as.numeric(orders)),
              m_fam = suppressWarnings(as.numeric(families)),
              m_gen = suppressWarnings(as.numeric(genera))), by = "row")

sapply(c("ord","fam","gen"), function(x) {
  a <- chk[[paste0("p_",x)]]; b <- chk[[paste0("m_",x)]]; ok <- is.finite(a) & is.finite(b)
  round(cor(a[ok], b[ok]), 3)
})
```

### Records to review

Informal entries (common names such as *fruit flies*, or the order alone) that
do not resolve to a species. No data is invented: they are listed verbatim for
review.

```{r review}
review <- long %>% filter(rank == "informal") %>%
  distinct(row, orchid, verbatim) %>%
  left_join(nonempty %>% transmute(row = .rowid, pollinators), by = "row") %>%
  arrange(row)

cat("Entries to review:", nrow(review), "\n")
head(review, 15)
```

## Pollinator taxonomy repository

A controlled vocabulary of the pollinators **present in this database** — the
order → family → genus hierarchy actually recorded, with how many orchid records
and pollinator species sit under each genus. This is the "repository of
plausible groups" that backs the drill-down menus (which only ever show present
taxa).

```{r taxonomy}
pollinator_taxonomy <- long %>%
  filter(!is.na(poll_genus)) %>%
  group_by(poll_group, poll_order, poll_family, poll_genus) %>%
  summarise(n_orchids = n_distinct(row),
            n_pollinator_spp = n_distinct(poll_taxon[rank == "taxon"]),
            .groups = "drop") %>%
  arrange(poll_group, poll_order, poll_family, poll_genus)

cat("Repository rows:", nrow(pollinator_taxonomy),
    "| orders:", n_distinct(pollinator_taxonomy$poll_order),
    "| families:", n_distinct(pollinator_taxonomy$poll_family),
    "| genera:", n_distinct(pollinator_taxonomy$poll_genus), "\n")
head(pollinator_taxonomy, 10)
```

### Names to review (likely typos)

We do not have an offline copy of the full world taxonomy, so instead of
validating against an external authority we look **within the database** for
names that are almost certainly the same: near-duplicate spellings, and genera
recorded under more than one family. Nothing is changed automatically — these
are flagged for a human to confirm (and, if real, added to the recode maps
above). `adist()` is edit (Levenshtein) distance.

```{r taxa-review}
# All within-vector near-duplicate pairs (edit distance <= max_d, shared prefix).
find_pairs <- function(v, max_d = 2, pre = 2, minlen = 5) {
  v <- sort(unique(v[!is.na(v) & nzchar(v)]))
  if (length(v) < 2) return(tibble(a = character(), b = character(), dist = integer()))
  m <- adist(v); idx <- which(upper.tri(m), arr.ind = TRUE)
  tibble(a = v[idx[,1]], b = v[idx[,2]], dist = as.integer(m[idx])) %>%
    filter(dist <= max_d, substr(a, 1, pre) == substr(b, 1, pre),
           pmin(nchar(a), nchar(b)) >= minlen)
}
gcount <- long %>% filter(!is.na(poll_genus)) %>% count(poll_genus, name = "n")

# (a) genus near-duplicates within the same family; keep the rarer as the typo
genus_dupes <- long %>% filter(!is.na(poll_genus), !is.na(poll_family)) %>%
  distinct(poll_family, poll_genus) %>%
  group_by(poll_family) %>% summarise(p = list(find_pairs(poll_genus)), .groups = "drop") %>%
  unnest(p) %>%
  left_join(gcount, by = c("a" = "poll_genus")) %>% rename(n_a = n) %>%
  left_join(gcount, by = c("b" = "poll_genus")) %>% rename(n_b = n) %>%
  transmute(issue = "Genus spelling (near-duplicate)", context = poll_family,
            name_1 = a, count_1 = n_a, name_2 = b, count_2 = n_b, distance = dist,
            suggestion = paste0("likely correct: ", ifelse(n_a >= n_b, a, b),
                                " · likely typo: ", ifelse(n_a >= n_b, b, a)))

# (b) a genus recorded under more than one family (usually a wrong family)
genus_multi <- long %>% filter(!is.na(poll_genus), !is.na(poll_family)) %>%
  distinct(poll_genus, poll_family) %>% add_count(poll_genus, name = "nf") %>%
  filter(nf > 1) %>% arrange(poll_genus, poll_family) %>%
  group_by(poll_genus) %>% summarise(families = paste(poll_family, collapse = "; "), .groups = "drop") %>%
  transmute(issue = "Genus under multiple families", context = families,
            name_1 = poll_genus, count_1 = NA_integer_, name_2 = NA_character_,
            count_2 = NA_integer_, distance = NA_integer_,
            suggestion = "verify family assignment")

# (c) family near-duplicates (strict: distance 1, since real families can be close)
family_dupes <- find_pairs(long$poll_family, max_d = 1, pre = 3, minlen = 6) %>%
  transmute(issue = "Family spelling (near-duplicate)", context = NA_character_,
            name_1 = a, count_1 = NA_integer_, name_2 = b, count_2 = NA_integer_,
            distance = dist, suggestion = "verify — may be distinct families")

taxa_review <- bind_rows(genus_dupes, genus_multi, family_dupes)
cat("Flagged for review:", nrow(taxa_review),
    " (genus near-dupes:", nrow(genus_dupes),
    ", genus multi-family:", nrow(genus_multi),
    ", family near-dupes:", nrow(family_dupes), ")\n")
head(genus_dupes, 12)
```

## Optional: validate against the GBIF backbone

The in-database checks above only compare names to each other. To catch typos
that happen to be the *only* spelling in the file — and to confirm which flagged
pairs are genuinely distinct genera (e.g. *Nomia* vs *Nomada*) — we validate the
present genera against the **GBIF taxonomic backbone**. This needs the `rgbif`
package and an internet connection (nothing else in this document does), so it
is written to run on your machine and to skip cleanly if either is missing.
Passing `kingdom = "Animalia"` avoids plant/animal homonyms (so *Agrostis* the
grass does not mask *Agrotis* the moth). Genera that do not match **exactly**
are written to `gbif_genus_review.csv` as strong typo/verification candidates.

```{r gbif-validate}
if (!requireNamespace("rgbif", quietly = TRUE)) {
  message("rgbif not installed - skipping GBIF check. Run install.packages('rgbif') to enable.")
} else {
  genera <- sort(unique(na.omit(pollinator_taxonomy$poll_genus)))
  res <- tryCatch(
    rgbif::name_backbone_checklist(data.frame(name = genera, kingdom = "Animalia")),
    error = function(e) { message("GBIF request failed: ", conditionMessage(e)); NULL })
  if (!is.null(res)) {
    gbif_genus_review <- dplyr::tibble(
      genus       = genera,
      gbif_name   = res$scientificName,
      gbif_rank   = res$rank,
      gbif_family = res$family,
      match_type  = res$matchType,
      confidence  = res$confidence) %>%
      dplyr::mutate(flag = is.na(match_type) | match_type != "EXACT") %>%
      dplyr::arrange(dplyr::desc(flag), genus)
    write.csv(gbif_genus_review, "gbif_genus_review.csv", row.names = FALSE)
    cat("GBIF checked", nrow(gbif_genus_review), "genera;",
        sum(gbif_genus_review$flag), "were not an exact match",
        "(written to gbif_genus_review.csv)\n")
    print(head(dplyr::filter(gbif_genus_review, flag), 15))
  }
}
```

## Exporting the enriched workbook

We save a **copy** with five sheets: `species` (original + wide columns),
`pollinators_long` (the long table), `needs_review`, `pollinator_taxonomy`
(the present order→family→genus repository) and `taxa_review` (likely typos).
The raw workbook is not modified.

```{r export}
salida <- "Pollination_List_pollinators_enriched.xlsx"
write.xlsx(
  list(species = species_out, pollinators_long = long, needs_review = review,
       pollinator_taxonomy = pollinator_taxonomy, taxa_review = taxa_review),
  salida, overwrite = TRUE)
cat("Written:", salida, "\n")
```

## Using it in Leaflet / Shiny

The new columns make the map **searchable by pollinator**. Two complementary
routes:

**In the Shiny app** (what you already have) — add a filter populated from the
parsed genera, alongside the existing subfamily filter:

```r
# UI: a searchable pollinator-genus selector
selectizeInput("gen_pol", "Pollinator genus:",
               choices = c("All", sort(unique(long$poll_genus))),
               options = list(placeholder = "type to search..."))

# server: keep the points whose poll_genera contains the chosen genus
df <- data_mapa
if (input$gen_pol != "All")
  df <- df %>% filter(str_detect(poll_genera, fixed(input$gen_pol)))
```

**On the static map** — search over the markers with `leaflet.extras`, and show
the clean list of taxa in the popup:

```r
library(leaflet.extras)
# include poll_taxa / poll_groups in popup_info, then:
mapa %>% addSearchFeatures(
  targetGroups = "orchids",
  options = searchFeaturesOptions(zoom = 6, openPopup = TRUE))
```

The `grp_bees … grp_lepid` flags are also ready for toggling layers by group,
and `n_taxa`, `n_genera`, etc. can size or color the points by pollinator
richness.
