---
title: "Cleaning and structuring the *pollinators* column"
subtitle: "Global Orchid Pollinators — preparing the data for search and Leaflet"
author: "Raymond L. Tremblay, Natalia Palou, Caleb Pacheco and Naan Lamoso"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-float: true
    theme: flatly
    df-print: paged
    code-fold: show
lang: en
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Goal

The **`pollinators`** column (column AF in the Excel) stores the pollinators as
free text, following a fairly consistent grammar:

```
ORDER:  Genus species, G. species2 (Family); Genus2 species (Family2).  ORDER2:  ...
```

That is: one or more **orders** in UPPERCASE followed by `:`; within each order,
a comma-separated list of taxa (often abbreviated after the first mention,
`Trigona melanocephala, T. minangkabau`), grouped into `(Family)`; families are
separated by `;` and orders by `.`.

The challenge we wanted to solve is twofold and seemingly contradictory: when a
record has **more than one pollinator** we need to be able to **separate** them
(to search by a specific genus or family) and at the same time **amalgamate**
them (so each point on the map can show a summary). The solution is to do a
**single parse** and derive **two views** from it:

1. A **long table** (`pollinators_long`): one row per *orchid × pollinator
   taxon*. This is the *separated* view, searchable at any rank (order, family,
   genus, species).
2. A set of **wide columns** (`poll_*`) joined back onto each orchid row. This
   is the *amalgamated* view, and it feeds the map popup and filters. They are
   generated **from** the long table, so the two never fall out of sync.

The original `pollinators` column is **left untouched**: it is the raw source.
All cleaning (including order and family typos) lives here, in the code — the
same philosophy `mapa_final.qmd` uses with the `wasp → wasps` recode.

***

## Packages

```{r libraries}
library(readxl)    # read the Excel
library(dplyr)     # data manipulation
library(tidyr)     # fill(), unnest()
library(stringr)   # regular expressions
library(purrr)     # map()
library(openxlsx)  # write the enriched workbook (multiple sheets)
```

## Loading the data

We read the `species` sheet of the master workbook and clean the column names,
just like the main analysis.

```{r load}
# The only line with the file name: if it is renamed, change it here.
# The name carries no date on purpose -- the master is updated in place, so the
# path stays valid and nothing downstream has to be edited again.
master_path <- "Pollination_List_MASTER.xlsx"

data <- read_excel(master_path, sheet = "species")
colnames(data) <- tolower(trimws(colnames(data)))

# Clean data-entry errors in `subfamily` BEFORE carrying it down. Orchidaceae
# has exactly five subfamilies; everything else here is an error:
#   "2"           -> Epidendroideae  (those rows are Bulbophyllum)
#   "Orchidaceae" -> Orchidoideae    (those rows are Sirindhornia)
#   "5" / "subfamily" -> NA          (junk header rows, genus == "genus")
subf_fix <- c("2" = "Epidendroideae", "Orchidaceae" = "Orchidoideae",
              "5" = NA_character_, "subfamily" = NA_character_)
sv <- trimws(as.character(data$subfamily))
data$subfamily <- ifelse(sv %in% names(subf_fix), unname(subf_fix[sv]), sv)

# Carry `genus` and `subfamily` down (they are written only once). This must
# run before any row filtering so the original row order stays intact.
fill_down <- function(df, col) {
  v <- trimws(as.character(df[[col]]))
  df[[col]] <- ifelse(v == "" | is.na(v), NA_character_, v)
  tidyr::fill(df, all_of(col), .direction = "down")
}
data <- data %>% fill_down("genus") %>% fill_down("subfamily")

# Full scientific name (genus + epithet) and a row id to rejoin on later.
data$orchid <- str_squish(paste(ifelse(is.na(data$genus), "", data$genus),
                                ifelse(is.na(data$species), "", data$species)))
data$.rowid <- seq_len(nrow(data))
```

## Species traits

The workbook also carries many one-hot trait columns (a `1` marks the trait,
blank means absent): growth form, climate zone, breeding system, and floral
rewards / deception strategy. We turn those scattered flags into a few
readable, searchable columns. Absent traits stay `NA` rather than "0", so
"unknown" and "recorded absent" are not confused.

```{r traits}
# TRUE only when a flag cell literally holds 1 / TRUE.
is_present <- function(x) tolower(trimws(as.character(x))) %in% c("1", "true")
flag <- function(name) if (name %in% names(data)) is_present(data[[name]]) else rep(FALSE, nrow(data))

# Combine several flag columns into one "A; B" label column (NA if none set).
combine_flags <- function(specs) {           # specs: c(Label = "column_name", ...)
  out <- rep("", nrow(data))
  for (k in seq_along(specs)) {
    on <- flag(unname(specs[k]))
    out <- ifelse(on, ifelse(out == "", names(specs)[k], paste(out, names(specs)[k], sep = "; ")), out)
  }
  ifelse(out == "", NA_character_, out)
}

data$growth_form   <- combine_flags(c(Epiphytic = "epi", Terrestrial = "terr",
                                      Rupicolous = "rupic", Climbing = "vine"))
data$climate       <- combine_flags(c(Tropical = "trop", Subtropical = "subtr",
                                      Temperate = "temp"))
data$compatibility <- combine_flags(c(`Self-incompatible` = "si", `Self-compatible` = "sc"))
data$breeding_system <- combine_flags(c(Chasmogamous = "chasmogamous",
                                       `Mixed mating` = "mixed mating",
                                       `Autonomous selfing/agamospermy` = "autonomous selfing/agagamospermy"))
data$reward        <- combine_flags(c(Nectar = "nectar", Pollen = "pollen",
                                     Fragrance = "fragr", `Oils/lipids` = "lipids",
                                     Pseudopollen = "trichomes/pseudo-pollen",
                                     `Sleeping site` = "sleep sites", Resins = "polyphenols resins"))
data$deception     <- combine_flags(c(`Food deception` = "food dec",
                                      `Sexual deception` = "sex dec",
                                      `Brood-site deception` = "brood site dec"))
data$mycoheterotrophic <- ifelse(flag("achlorophyllous"), "Mycoheterotrophic", NA_character_)

# How many records carry each derived trait (a quick fill check).
sapply(c("growth_form","climate","compatibility","breeding_system",
         "reward","deception","mycoheterotrophic"),
       function(c) sum(!is.na(data[[c]])))
```

## Recode maps (typos)

Orders have ~10 real values but ~20 different spellings; families have similar
typos. We fix them with two small dictionaries that are easy to extend as new
variants appear.

```{r recodes}
order_recode <- c(
  DIPERA = "DIPTERA", HYEMNOPTERA = "HYMENOPTERA", HHYMENOPTERA = "HYMENOPTERA",
  HYMENOTERA = "HYMENOPTERA", HYMENOTPERA = "HYMENOPTERA",
  LEPIODOPTERA = "LEPIDOPTERA", LEPTIDOPTERA = "LEPIDOPTERA", HEMPTERIA = "HEMIPTERA")

family_recode <- c(
  Andrinidae = "Andrenidae", Aipdae = "Apidae", Mellitidae = "Melittidae",
  Anthomyidae = "Anthomyiidae", Choropidae = "Chloropidae", Hesperidae = "Hesperiidae",
  Tephrididae = "Tephritidae", Empidae = "Empididae", Papillionidae = "Papilionidae",
  Papilioidae = "Papilionidae", Cramabidae = "Crambidae", Nymphaidae = "Nymphalidae",
  Bombylidae = "Bombyliidae", Calliphridae = "Calliphoridae", Syriphidae = "Syrphidae",
  Thynnindae = "Thynnidae", Thinnidae = "Thynnidae", Thyniidae = "Thynnidae",
  Thiniidae = "Thynnidae", Pompiliidae = "Pompilidae", Pompylidae = "Pompilidae",
  Scatophagidae = "Scathophagidae", Chrysidae = "Chrysididae", Scolidae = "Scoliidae",
  Cullicidae = "Culicidae", Mycetophiilidae = "Mycetophilidae", Saturnidae = "Saturniidae",
  Tabinidae = "Tabanidae", Tenthredionidae = "Tenthredinidae", Erabidae = "Erebidae",
  Zosteropsidae = "Zosteropidae", Ullidiidae = "Ulidiidae", Apidea = "Apidae")

fix_order  <- function(o) { o <- toupper(o); ifelse(o %in% names(order_recode),
                                                    unname(order_recode[o]), o) }
fix_family <- function(x) ifelse(!is.na(x) & x %in% names(family_recode),
                                 unname(family_recode[x]), x)

# Coarse group, using full taxonomic names (no contractions). Whole orders
# keep the order name; HYMENOPTERA is split into Bees vs. Wasps, because that
# distinction matters for pollination ecology.
bee_fams <- c("Apidae","Halictidae","Andrenidae","Megachilidae","Colletidae",
              "Melittidae","Stenotritidae")

# Family is the primary signal, but many source papers name the bee without its
# family ("HYMENOPTERA: Bombus cf. albopleuralis"), or misspell it. Falling
# straight through to "Wasps" in that case mislabels obvious bees, so we check
# the GENUS as a second signal before giving up. Extend this list as needed.
bee_genera <- c("Apis","Bombus","Xylocopa","Euglossa","Eulaema","Eufriesea",
                "Exaerete","Melipona","Trigona","Tetragonisca","Nannotrigona",
                "Scaptotrigona","Centris","Ceratina","Tetralonia","Tetraloniella",
                "Eucera","Anthophora","Andrena","Nomada","Lasioglossum","Halictus",
                "Augochlora","Augochlorella","Augochloropsis","Megachile","Osmia",
                "Colletes","Hylaeus","Melissodes","Peponapis","Xylocopinae")

group_from <- function(order, family, genus = NA_character_) {
  order <- toupper(order); if (is.na(order)) return(NA_character_)
  switch(order,
    DIPTERA = "Diptera", COLEOPTERA = "Coleoptera", LEPIDOPTERA = "Lepidoptera",
    HEMIPTERA = "Hemiptera", THYSANOPTERA = "Thysanoptera", ORTHOPTERA = "Orthoptera",
    ARANEAE = "Arachnida", ARACHNIDA = "Arachnida", AVES = "Aves",
    HYMENOPTERA =
      if (!is.na(family) && family %in% bee_fams) "Bees"
      else if (!is.na(genus)  && genus  %in% bee_genera) "Bees"
      # Neither family nor genus identifies it. "Wasps" would be a guess, so say
      # so plainly instead -- these rows are the ones worth checking by hand.
      else if (is.na(family) && is.na(genus)) "Hymenoptera (unassigned)"
      else "Wasps",
    str_to_title(tolower(order)))   # any other order -> proper-case name
}
```

## Genus spellings confirmed by GBIF

`order_recode` and `family_recode` above fix typos we identified by eye. This map
does the same at genus rank, but every entry was confirmed against the GBIF
backbone rather than judged here: in each case GBIF resolved both spellings to
the same accepted name, or placed one spelling and could not place the other at
all. No inference was involved, so these are applied automatically.

Corrections that depended on *reasoning* — the family-consistency rule, which
assumes the family a record was published under is correct — are deliberately
**not** here. They are listed in the report under "Automatic corrections that
overrode a real name" and are held until a specialist confirms them.

Suprageneric and informal names are also excluded on purpose. `Anthophorid`,
`Halictid` and `Meliponids` mean "a bee of that family"; GBIF will happily
resolve them to *Anthophora*, *Halictus* and so on, but recording them as a
genus would assert precision the source never gave.

```{r genus-recode}
genus_recode <- c(
  "Adscrita" = "Adscita", "Aeroptetes" = "Aeropetes",
  "Agaostemon" = "Agapostemon", "Agopostemon" = "Agapostemon",
  "Alllodape" = "Allodape", "Amastas" = "Amastus",
  "Amazalia" = "Amazilia", "Amazilla" = "Amazilia",
  "Amegella" = "Amegilla", "Andrenia" = "Andrena",
  "Apamaea" = "Apamea", "Atricelaphinis" = "Atrichelaphinis",
  "Atrichelaphinus" = "Atrichelaphinis", "Auglochloropsis" = "Augochloropsis",
  "Augochora" = "Augochlora", "Basiotha" = "Basiothia",
  "Bombu" = "Bombus", "Campsomesiella" = "Campsomeriella",
  "Chaliocodoma" = "Chalicodoma", "Cheilosa" = "Cheilosia",
  "Chysura" = "Chrysura", "Cochinella" = "Coccinella",
  "Costacovexa" = "Costaconvexa", "Cytothyrea" = "Cyrtothyrea",
  "Dolichvespula" = "Dolichovespula", "Entepheria" = "Entephria",
  "Entephtia" = "Entephria", "Entriphria" = "Entephria",
  "Eparygyreus" = "Epargyreus", "Erisalis" = "Eristalis",
  "Eufrisea" = "Eufriesea", "Eugloaa" = "Euglossa",
  "Euremea" = "Eurema", "Euriesea" = "Eufriesea",
  "Exarete" = "Exaerete", "Exomalopsiis" = "Exomalopsis",
  "Gnophuos" = "Gnophos", "Halictis" = "Halictus",
  "Heliothus" = "Heliothis", "Helphilus" = "Helophilus",
  "Hemipepis" = "Hemipepsis", "Ischioden" = "Ischiodon",
  "Ishiodon" = "Ischiodon", "Leptothoras" = "Leptothorax",
  "Maroglossum" = "Macroglossum", "Megahile" = "Megachile",
  "Oedemeria" = "Oedemera", "Parabolybia" = "Parapolybia",
  "Parapoolybia" = "Parapolybia", "Paratetrapaedia" = "Paratetrapedia",
  "Peritichia" = "Peritrichia", "Procoeca" = "Prosoeca",
  "Scatophagia" = "Scatophaga", "Syrrita" = "Syritta",
  "Tachinia" = "Tachina", "Tetrpedia" = "Tetrapedia",
  "Thynoides" = "Thynnoides", "Trchoplusia" = "Trichoplusia",
  "Xyphanes" = "Xylophanes", "Zasipilothynnus" = "Zaspilothynnus")

fix_genus <- function(x) ifelse(!is.na(x) & x %in% names(genus_recode),
                               unname(genus_recode[x]), x)
cat("Genus spellings corrected from GBIF:", length(genus_recode), "\n")
```

## Parsing function

`parse_poll()` takes **one** `pollinators` cell and returns a table with one
row per taxon: order, family, genus, epithet, full binomial, the original text
(`verbatim`), the rank reached (`taxon` / `genus` / `informal`) and the coarse
group.

```{r parser}
# Common English words that are NOT scientific names (they prevent false
# genera/species such as "Stingless bees" or "fruit flies").
COMMON <- c("stingless","fruit","flies","bees","wasps","moths","beetles",
            "birds","ants","hawk","small","large")

parse_poll <- function(s) {
  if (is.na(s) || !nzchar(trimws(s))) return(tibble())
  s <- str_squish(s)
  # ORDER markers: an UPPERCASE word (>=4 letters) followed by ':'
  m <- str_locate_all(s, "[A-Z]{4,}\\s*:")[[1]]
  if (nrow(m) == 0)
    return(tibble(order = NA_character_, family = NA_character_, genus = NA_character_,
                  epithet = NA_character_, taxon = NA_character_, verbatim = s,
                  rank = "informal"))
  starts <- m[,1]; endc <- m[,2]
  onames <- trimws(str_replace(str_sub(s, starts, endc), ":$", ""))
  bend <- c(starts[-1] - 1, nchar(s)); last_genus <- NA_character_; out <- list()

  for (i in seq_along(onames)) {
    order <- fix_order(onames[i])
    body  <- str_squish(str_sub(s, endc[i] + 1, bend[i]))
    if (!nzchar(body)) next
    # Each segment = one family. Cut after a ')' followed by ';' or '.', and
    # also on lone ';' (taxa with no family in parentheses).
    segs <- unlist(str_split(str_split(body, "(?<=\\))\\s*[;.]+\\s*")[[1]], ";"))
    for (seg in segs) {
      seg <- str_squish(str_replace(seg, "[\\.\\s]+$", ""))
      if (!nzchar(seg)) next
      fam <- NA_character_
      par <- str_match(seg, "\\(([^)]*)\\)\\s*$")          # last parenthesis
      if (!is.na(par[1,2])) fam <- fix_family(str_extract(par[1,2], "[A-Z][a-z]+dae"))
      taxa <- str_squish(str_replace(seg, "\\([^)]*\\)\\s*$", ""))  # drop the family

      for (tok in str_split(taxa, ",")[[1]]) {
        tok <- str_squish(tok); if (!nzchar(tok)) next
        genus <- NA_character_; epithet <- NA_character_; rank <- "taxon"
        if (str_detect(tok, "^[A-Z]\\.\\s*")) {            # abbreviated genus "B. ..."
          rest  <- str_squish(str_replace(tok, "^[A-Z]\\.\\s*", ""))
          genus <- last_genus                              # expand with the last full genus
          epithet <- str_extract(rest, "^[a-z][a-z-]+")
        } else if (str_detect(tok, "^[A-Z][a-z]+")) {      # full genus
          genus <- fix_genus(str_extract(tok, "^[A-Z][a-z]+")); last_genus <- genus
          rest  <- str_squish(str_replace(tok, "^[A-Z][a-z]+\\s*", ""))
          epithet <- str_extract(rest, "^[a-z][a-z-]+")
        } else rank <- "informal"
        # sp./spp./aff./cf. -> genus level
        if (!is.na(epithet) && epithet %in% c("sp","spp","aff","cf","near")) epithet <- NA_character_
        if (!is.na(epithet) && epithet %in% COMMON) { epithet <- NA_character_; rank <- "informal" }
        if (!is.na(genus)   && tolower(genus) %in% COMMON) { genus <- NA_character_; rank <- "informal" }
        if (is.na(genus) && rank == "taxon") rank <- "informal"
        if (rank == "taxon" && is.na(epithet)) rank <- "genus"
        taxon <- if (rank == "taxon") paste(genus, epithet) else if (rank == "genus") genus else NA_character_
        out[[length(out) + 1]] <- tibble(order = order, family = fam, genus = genus,
                                         epithet = epithet, taxon = taxon,
                                         verbatim = tok, rank = rank)
      }
    }
  }
  if (length(out) == 0)
    return(tibble(order = NA, family = NA, genus = NA, epithet = NA, taxon = NA,
                  verbatim = s, rank = "informal"))
  bind_rows(out) %>% mutate(group = pmap_chr(list(order, family, genus), group_from))
}
```

## Long table (the *separated* view)

```{r long}
nonempty <- data %>% filter(!is.na(pollinators) & nzchar(trimws(pollinators)))

long <- nonempty %>%
  mutate(pp = map(pollinators, parse_poll)) %>%
  transmute(row = .rowid, orchid, subfamily, pollinator_type, pollinators, pp) %>%
  unnest(pp) %>%
  transmute(row, orchid, subfamily, pollinator_type,
            poll_order = order, poll_family = family, poll_genus = genus,
            poll_epithet = epithet, poll_taxon = taxon, poll_group = group,
            rank, verbatim)

head(long, 12)
```

## Wide columns (the *amalgamated* view)

These are built by grouping the long table by orchid record, so they are a
faithful summary of the separated data.

```{r wide}
j <- function(x) paste(sort(unique(x[!is.na(x) & nzchar(x)])), collapse = "; ")

wide <- long %>%
  group_by(row) %>%
  summarise(
    poll_orders   = j(poll_order),
    poll_families = j(poll_family),
    poll_genera   = j(poll_genus),
    poll_taxa     = j(poll_taxon),
    poll_groups   = j(poll_group),
    n_orders   = n_distinct(poll_order[!is.na(poll_order)]),
    n_families = n_distinct(poll_family[!is.na(poll_family)]),
    n_genera   = n_distinct(poll_genus[!is.na(poll_genus)]),
    n_taxa     = n_distinct(poll_taxon[rank == "taxon"]),
    grp_bees    = any(poll_group == "Bees"),
    grp_diptera = any(poll_group == "Diptera"),
    grp_wasps   = any(poll_group == "Wasps"),
    grp_coleo   = any(poll_group == "Coleoptera"),
    grp_lepid   = any(poll_group == "Lepidoptera"),
    .groups = "drop")

# Enriched `species` sheet = original columns + wide columns.
species_out <- data %>%
  left_join(wide, by = c(".rowid" = "row")) %>%
  select(-.rowid, -orchid)

head(wide, 8)
```

## Quality control

```{r qc}
cat("Long rows (taxa):", nrow(long), "\n")
cat("Orchid records with pollinators:", nrow(wide), "\n")
cat("Orders:", n_distinct(long$poll_order, na.rm = TRUE),
    "| Families:", n_distinct(long$poll_family, na.rm = TRUE),
    "| Genera:", n_distinct(long$poll_genus, na.rm = TRUE),
    "| Full binomials:", n_distinct(long$poll_taxon[long$rank == "taxon"]), "\n\n")

long %>% count(rank, name = "rows") %>% knitr::kable(caption = "Rows by rank reached")
long %>% count(poll_group, name = "rows") %>% arrange(desc(rows)) %>%
  knitr::kable(caption = "Rows by coarse group")
```

We compare our counts with the columns that were already tallied by hand in the
Excel (`orders`, `families`, `genera`): a high correlation confirms the parse
reproduces the original criterion.

```{r qc-cross}
chk <- long %>% group_by(row) %>%
  summarise(p_ord = n_distinct(poll_order, na.rm = TRUE),
            p_fam = n_distinct(poll_family, na.rm = TRUE),
            p_gen = n_distinct(poll_genus, na.rm = TRUE), .groups = "drop") %>%
  left_join(nonempty %>% transmute(row = .rowid,
              m_ord = suppressWarnings(as.numeric(orders)),
              m_fam = suppressWarnings(as.numeric(families)),
              m_gen = suppressWarnings(as.numeric(genera))), by = "row")

sapply(c("ord","fam","gen"), function(x) {
  a <- chk[[paste0("p_",x)]]; b <- chk[[paste0("m_",x)]]; ok <- is.finite(a) & is.finite(b)
  round(cor(a[ok], b[ok]), 3)
})
```

### Records to review

Informal entries (common names such as *fruit flies*, or the order alone) that
do not resolve to a species. No data is invented: they are listed verbatim for
review.

```{r review}
review <- long %>% filter(rank == "informal") %>%
  distinct(row, orchid, verbatim) %>%
  left_join(nonempty %>% transmute(row = .rowid, pollinators), by = "row") %>%
  arrange(row)

cat("Entries to review:", nrow(review), "\n")
head(review, 15)
```

## Pollinator taxonomy repository

A controlled vocabulary of the pollinators **present in this database** — the
order → family → genus hierarchy actually recorded, with how many orchid records
and pollinator species sit under each genus. This is the "repository of
plausible groups" that backs the drill-down menus (which only ever show present
taxa).

```{r taxonomy}
pollinator_taxonomy <- long %>%
  filter(!is.na(poll_genus)) %>%
  group_by(poll_group, poll_order, poll_family, poll_genus) %>%
  summarise(n_orchids = n_distinct(row),
            n_pollinator_spp = n_distinct(poll_taxon[rank == "taxon"]),
            .groups = "drop") %>%
  arrange(poll_group, poll_order, poll_family, poll_genus)

cat("Repository rows:", nrow(pollinator_taxonomy),
    "| orders:", n_distinct(pollinator_taxonomy$poll_order),
    "| families:", n_distinct(pollinator_taxonomy$poll_family),
    "| genera:", n_distinct(pollinator_taxonomy$poll_genus), "\n")
head(pollinator_taxonomy, 10)
```

### Names to review (likely typos)

First we look **within the database** for names that are almost certainly the
same: near-duplicate spellings, and genera recorded under more than one family.
This pass is deliberately generous — it flags anything that *might* be an error,
and it has no external authority to consult, so the list below is a starting
point rather than a verdict. `adist()` is edit (Levenshtein) distance.

Most of these are settled a few sections down, once the names have been checked
against the GBIF backbone; see **What the GBIF check settles**. Nothing is
changed automatically at any stage.

```{r taxa-review}
# All within-vector near-duplicate pairs (edit distance <= max_d, shared prefix).
find_pairs <- function(v, max_d = 2, pre = 2, minlen = 5) {
  v <- sort(unique(v[!is.na(v) & nzchar(v)]))
  if (length(v) < 2) return(tibble(a = character(), b = character(), dist = integer()))
  m <- adist(v); idx <- which(upper.tri(m), arr.ind = TRUE)
  tibble(a = v[idx[,1]], b = v[idx[,2]], dist = as.integer(m[idx])) %>%
    filter(dist <= max_d, substr(a, 1, pre) == substr(b, 1, pre),
           pmin(nchar(a), nchar(b)) >= minlen)
}
gcount <- long %>% filter(!is.na(poll_genus)) %>% count(poll_genus, name = "n")

# (a) genus near-duplicates within the same family; keep the rarer as the typo
genus_dupes <- long %>% filter(!is.na(poll_genus), !is.na(poll_family)) %>%
  distinct(poll_family, poll_genus) %>%
  group_by(poll_family) %>% summarise(p = list(find_pairs(poll_genus)), .groups = "drop") %>%
  unnest(p) %>%
  left_join(gcount, by = c("a" = "poll_genus")) %>% rename(n_a = n) %>%
  left_join(gcount, by = c("b" = "poll_genus")) %>% rename(n_b = n) %>%
  transmute(issue = "Genus spelling (near-duplicate)", context = poll_family,
            name_1 = a, count_1 = n_a, name_2 = b, count_2 = n_b, distance = dist,
            suggestion = paste0("likely correct: ", ifelse(n_a >= n_b, a, b),
                                " · likely typo: ", ifelse(n_a >= n_b, b, a)))

# (b) a genus recorded under more than one family (usually a wrong family)
genus_multi <- long %>% filter(!is.na(poll_genus), !is.na(poll_family)) %>%
  distinct(poll_genus, poll_family) %>% add_count(poll_genus, name = "nf") %>%
  filter(nf > 1) %>% arrange(poll_genus, poll_family) %>%
  group_by(poll_genus) %>% summarise(families = paste(poll_family, collapse = "; "), .groups = "drop") %>%
  transmute(issue = "Genus under multiple families", context = families,
            name_1 = poll_genus, count_1 = NA_integer_, name_2 = NA_character_,
            count_2 = NA_integer_, distance = NA_integer_,
            suggestion = "verify family assignment")

# (c) family near-duplicates (strict: distance 1, since real families can be close)
family_dupes <- find_pairs(long$poll_family, max_d = 1, pre = 3, minlen = 6) %>%
  transmute(issue = "Family spelling (near-duplicate)", context = NA_character_,
            name_1 = a, count_1 = NA_integer_, name_2 = b, count_2 = NA_integer_,
            distance = dist, suggestion = "verify — may be distinct families")

taxa_review <- bind_rows(genus_dupes, genus_multi, family_dupes)
cat("Flagged for review:", nrow(taxa_review),
    " (genus near-dupes:", nrow(genus_dupes),
    ", genus multi-family:", nrow(genus_multi),
    ", family near-dupes:", nrow(family_dupes), ")\n")
head(genus_dupes, 12)
```

## Optional: validate against the GBIF backbone

The in-database checks above only compare names to each other. To catch typos
that happen to be the *only* spelling in the file — and to confirm which flagged
pairs are genuinely distinct genera (e.g. *Nomia* vs *Nomada*) — we validate the
present genera against the **GBIF taxonomic backbone**. This needs the `rgbif`
package and an internet connection (nothing else in this document does), so it
is written to run on your machine and to report loudly if either is missing.

GBIF will not choose between equally-good homonyms — it returns their lowest
common ancestor instead. *Anthophora* matches the bee genus (Apidae) and a
stony coral (Pocilloporidae) at identical confidence, so the plain matcher
answers "Animalia" and the name appears unresolvable. We therefore query each
genus verbosely and, when the direct match lands above genus rank, take the best
genus-rank candidate belonging to a class that could actually pollinate an
orchid. Those cases are labelled `disambiguated` rather than `direct`, so the
inference is visible instead of buried. Only rows marked `unresolved` need a
human decision; they are written to `gbif_genus_review.csv`.

```{r gbif-validate, message = TRUE, warning = TRUE}
# NOTE: report with cat(), not message(). `opts_chunk$set(message = FALSE)` at
# the top of this document hides message() output, which once let this chunk
# skip in silence and leave a stale gbif_genus_review.csv on disk.
if (!requireNamespace("rgbif", quietly = TRUE)) {
  cat("SKIPPED: rgbif is not installed. Run install.packages('rgbif') to enable",
      "the GBIF check; gbif_genus_review.csv has NOT been updated.\n")
} else {
  genera <- sort(unique(na.omit(pollinator_taxonomy$poll_genus)))

  # Classes a pollinator of an orchid can plausibly belong to. Used only to
  # break homonym ties -- see below.
  poll_classes <- c("Insecta", "Arachnida", "Aves", "Mammalia", "Reptilia")

  # WHY THIS IS DONE ONE NAME AT A TIME, AND VERBOSELY.
  #
  # GBIF's matcher refuses to guess between equally-good homonyms: it returns
  # the lowest common ancestor instead. `Anthophora` matches both the bee genus
  # (Apidae, Insecta) and a stony coral (Pocilloporidae, Anthozoa), each at
  # confidence 92, so the plain matcher answers "Animalia" and the name looks
  # unresolvable. It is not -- the candidates are in the verbose response.
  # Passing `rank`/`kingdom` does not help; the tie is what stops it.
  #
  # So: take the direct match when it lands at genus rank, and otherwise pick
  # the best genus-rank alternative that sits in a class capable of pollinating
  # an orchid. A coral is not a pollinator; that is the whole of the judgement
  # being applied here, and every such case is labelled "disambiguated" in the
  # output so it can be audited rather than trusted.
  resolve_genus <- function(g) {
    out <- dplyr::tibble(genus = g, gbif_name = NA_character_,
                         gbif_rank = NA_character_, gbif_family = NA_character_,
                         gbif_class = NA_character_, match_type = NA_character_,
                         confidence = NA_integer_, resolution = "unresolved")
    nb <- tryCatch(rgbif::name_backbone_verbose(name = g, kingdom = "Animalia"),
                   error = function(e) NULL)
    if (is.null(nb)) return(out)

    d <- nb$data
    if (!is.null(d) && nrow(d) > 0 && !is.na(d$rank[1]) && d$rank[1] == "GENUS") {
      out$gbif_name   <- d$scientificName[1]
      out$gbif_rank   <- "GENUS"
      out$gbif_family <- if ("family" %in% names(d)) d$family[1] else NA_character_
      out$gbif_class  <- if ("class"  %in% names(d)) d$class[1]  else NA_character_
      out$match_type  <- d$matchType[1]
      out$confidence  <- d$confidence[1]
      out$resolution  <- "direct"
      return(out)
    }

    a <- nb$alternatives
    if (!is.null(a) && nrow(a) > 0 && all(c("rank", "class") %in% names(a))) {
      cand <- a %>%
        dplyr::filter(rank == "GENUS", class %in% poll_classes) %>%
        # an accepted name beats a synonym; then highest confidence
        dplyr::mutate(.accepted = as.integer(status == "ACCEPTED")) %>%
        dplyr::arrange(dplyr::desc(.accepted), dplyr::desc(confidence))
      if (nrow(cand) > 0) {
        out$gbif_name   <- cand$scientificName[1]
        out$gbif_rank   <- "GENUS"
        out$gbif_family <- cand$family[1]
        out$gbif_class  <- cand$class[1]
        out$match_type  <- "DISAMBIGUATED"
        out$confidence  <- cand$confidence[1]
        # more than one plausible pollinator genus == still worth a human look
        out$resolution  <- if (dplyr::n_distinct(cand$family) > 1)
          "disambiguated (several candidate families)" else "disambiguated"
      }
    }
    out
  }

  cat("Querying GBIF for", length(genera), "genera, one at a time.",
      "This takes a few minutes.\n")
  gbif_genus_review <- purrr::map_dfr(genera, resolve_genus) %>%
    dplyr::mutate(flag = resolution != "direct") %>%
    dplyr::arrange(match(resolution, c("unresolved",
                                       "disambiguated (several candidate families)",
                                       "disambiguated", "direct")), genus)

  write.csv(gbif_genus_review, "gbif_genus_review.csv", row.names = FALSE)
  cat("\nGBIF resolution of", nrow(gbif_genus_review), "genera:\n")
  print(table(gbif_genus_review$resolution))
  cat("\nWritten to gbif_genus_review.csv.",
      "Only the 'unresolved' rows need a human decision.\n\n")
  print(head(dplyr::filter(gbif_genus_review, resolution != "direct"), 15))
}
```

### What the genus corrections actually changed

```{r genus-recode-effect, message = TRUE}
applied <- long %>%
  dplyr::filter(!is.na(poll_genus)) %>%
  dplyr::count(poll_genus, name = "rows_now") %>%
  dplyr::filter(poll_genus %in% unname(genus_recode))
changed <- tibble::tibble(
  wrong_spelling = names(genus_recode),
  corrected_to   = unname(genus_recode))
write.csv(changed, "genus_corrections_changelog.csv", row.names = FALSE)
cat("Applied", nrow(changed), "GBIF-confirmed genus spellings;",
    "written to genus_corrections_changelog.csv\n")
knitr::kable(changed, caption = "Genus spellings corrected automatically, each confirmed against GBIF.")
```

## What the GBIF check settles

The flagged list above is deliberately generous — it is looking for anything that
*might* be an error. Most of it is not a judgement call once the names are put to
an external authority. Here we join the GBIF result back onto `taxa_review` and
split it in two: what is settled, and what still needs a taxonomist.

A GBIF match is accepted only when the family it returns **agrees with the family
the record was published under**. That second condition matters more than it
looks. Several typos happen to match a real animal elsewhere in the tree —
`Megasella` resolves to a brachiopod, `Scaria` to a grasshopper, `Bombis` to a
moth — and without the family check those would be waved through as valid names.

The rule has a known weakness, stated here because it is invisible otherwise: it
assumes the recorded family is correct. Where an orchid genuinely has an
unexpected visitor, this would quietly correct it away. Rows where a rejected
spelling was itself a real genus are therefore listed separately below, so the
inference is visible rather than buried.

```{r taxa-review-status, message = TRUE}
if (!exists("gbif_genus_review")) {
  cat("No GBIF results in this session — every flagged name is still open.\n")
  taxa_review$gbif_status   <- NA_character_
  taxa_review$gbif_evidence <- NA_character_
} else {
  gb <- gbif_genus_review %>%
    dplyr::transmute(name = genus,
                     gen  = ifelse(is.na(gbif_rank), NA_character_, sub(" .*$", "", gbif_name)),
                     fam  = gbif_family)
  look <- function(x, col) { i <- match(x, gb$name); ifelse(is.na(i), NA_character_, gb[[col]][i]) }

  # Every family a genus is ACTUALLY recorded under, across the whole database.
  #
  # This must not be taken from the flagged pair's `context`, which is only the
  # single family in which the two names happened to collide. `Nomada` appears
  # under Apidae 19 times and Halictidae once; judged against Halictidae alone it
  # looks like a misspelling of `Nomia` and all 20 records would have been
  # renamed. Both are real bee genera.
  fam_of <- long %>%
    dplyr::filter(!is.na(poll_genus), !is.na(poll_family)) %>%
    dplyr::distinct(poll_genus, poll_family)
  recorded_families <- function(g) fam_of$poll_family[fam_of$poll_genus == g]

  classify <- function(issue, context, name_1, count_1, name_2, count_2) {
    if (identical(issue, "Genus spelling (near-duplicate)")) {
      n1 <- ifelse(is.na(count_1), 0, count_1); n2 <- ifelse(is.na(count_2), 0, count_2)
      common <- if (n1 >= n2) name_1 else name_2
      rare   <- if (n1 >= n2) name_2 else name_1
      kC <- look(common, "gen"); kR <- look(rare, "gen")
      fC <- look(common, "fam"); fR <- look(rare, "fam")
      if (!is.na(kC) && !is.na(kR) && kC == kR)
        return(c("Resolved", paste0("both spellings resolve to ", kC)))
      if (!is.na(kC) && is.na(kR))
        return(c("Resolved", paste0("'", common, "' = ", kC, " (", fC, "); '", rare, "' matches no genus")))
      if (is.na(kC) && !is.na(kR))
        return(c("Resolved", paste0("'", rare, "' = ", kR, " (", fR, "); '", common, "' matches no genus")))
      if (!is.na(kC) && !is.na(kR)) {
        okC <- !is.na(fC) && fC %in% recorded_families(common)
        okR <- !is.na(fR) && fR %in% recorded_families(rare)
        if (okC && !okR)
          return(c("Resolved", paste0("'", rare, "' matches ", kR, " in ", fR,
                                      ", a family it is never recorded under — fuzzy collision")))
        if (okR && !okC)
          return(c("Resolved", paste0("'", common, "' matches ", kC, " in ", fC,
                                      ", a family it is never recorded under")))
        return(c("Needs review",
                 paste0("both are real genera consistent with their own records: ",
                        kC, " (", fC, ") vs ", kR, " (", fR, ")")))
      }
      return(c("Needs review", "GBIF places neither spelling at genus rank"))
    }
    if (identical(issue, "Genus under multiple families")) {
      k <- look(name_1, "gen"); f <- look(name_1, "fam")
      rec <- trimws(unlist(strsplit(ifelse(is.na(context), "", context), ";")))
      if (is.na(k)) return(c("Needs review", "GBIF places this name at no genus rank"))
      if (!is.na(f) && f %in% rec)
        return(c("Resolved", paste0("GBIF: ", f, "; the other families in our records are errors")))
      return(c("Needs review", paste0("GBIF: ", f, ", which is not among the families our records use")))
    }
    c("Needs review", "family-level name — the GBIF genus check does not apply")
  }

  st <- purrr::pmap(list(taxa_review$issue, taxa_review$context, taxa_review$name_1,
                         taxa_review$count_1, taxa_review$name_2, taxa_review$count_2),
                    classify)
  taxa_review$gbif_status   <- vapply(st, `[`, character(1), 1)
  taxa_review$gbif_evidence <- vapply(st, `[`, character(1), 2)

  cat("Of", nrow(taxa_review), "flagged names, GBIF settles",
      sum(taxa_review$gbif_status == "Resolved"), "and leaves",
      sum(taxa_review$gbif_status == "Needs review"), "for a taxonomist.\n")
}
```

### Still unresolved — these need a specialist

```{r taxa-unresolved}
taxa_unresolved <- taxa_review %>% dplyr::filter(gbif_status == "Needs review")
taxa_unresolved %>%
  dplyr::select(issue, context, name_1, count_1, name_2, count_2, gbif_evidence) %>%
  knitr::kable(caption = paste0("The ", nrow(taxa_unresolved),
    " flagged names GBIF cannot settle. Everything else in `taxa_review` is resolved."))
```

### Automatic corrections that overrode a real name

Each row below is a spelling we treated as an error even though it matches a real
genus somewhere else. Most are plainly fuzzy collisions, but "plainly" is doing
work that only a specialist can check, so they are surfaced rather than assumed.

```{r taxa-overrides}
if (exists("gbif_genus_review")) {
  overrides <- taxa_review %>%
    dplyr::filter(issue == "Genus spelling (near-duplicate)", gbif_status == "Resolved",
                  grepl("fuzzy collision|outside the recorded family", gbif_evidence)) %>%
    dplyr::transmute(recorded_family = context, name_1, count_1, name_2, count_2,
                     what_we_concluded = gbif_evidence)
  knitr::kable(overrides, caption = paste0(nrow(overrides),
    " corrections that discarded a spelling which is itself a valid genus elsewhere."))
}
```

## Exporting the enriched workbook

We save a **copy** with five sheets: `species` (original + wide columns),
`pollinators_long` (the long table), `needs_review`, `pollinator_taxonomy`
(the present order→family→genus repository) and `taxa_review` (flagged names, each
carrying the `gbif_status` and `gbif_evidence` columns added above).
The raw workbook is not modified.

```{r export}
salida <- "Pollination_List_pollinators_enriched.xlsx"
write.xlsx(
  list(species = species_out, pollinators_long = long, needs_review = review,
       pollinator_taxonomy = pollinator_taxonomy, taxa_review = taxa_review),
  salida, overwrite = TRUE)
cat("Written:", salida, "\n")
```

## Using it in Leaflet / Shiny

The new columns make the map **searchable by pollinator**. Two complementary
routes:

**In the Shiny app** (what you already have) — add a filter populated from the
parsed genera, alongside the existing subfamily filter:

```r
# UI: a searchable pollinator-genus selector
selectizeInput("gen_pol", "Pollinator genus:",
               choices = c("All", sort(unique(long$poll_genus))),
               options = list(placeholder = "type to search..."))

# server: keep the points whose poll_genera contains the chosen genus
df <- data_mapa
if (input$gen_pol != "All")
  df <- df %>% filter(str_detect(poll_genera, fixed(input$gen_pol)))
```

**On the static map** — search over the markers with `leaflet.extras`, and show
the clean list of taxa in the popup:

```r
library(leaflet.extras)
# include poll_taxa / poll_groups in popup_info, then:
mapa %>% addSearchFeatures(
  targetGroups = "orchids",
  options = searchFeaturesOptions(zoom = 6, openPopup = TRUE))
```

The `grp_bees … grp_lepid` flags are also ready for toggling layers by group,
and `n_taxa`, `n_genera`, etc. can size or color the points by pollinator
richness.
